---
title: "Clinical Data Preparation"
output:
  html_document:
    df_print: paged
---


```{r include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
#source("Functions.R")
library(readxl)
library(paletteer)
```

## 0. Read Files

```{r echo=TRUE, warning=FALSE}

abpm_file <- read_excel("C:/Users/rosa/Documents/4CStudy/ClinicalData/abpm_dataset_4C_for_OLINK_with_SDS_values_by_lms_method_for_abpm_measures_(JH-24-NOV-2024).xlsx")
investigation_file <- read_excel("C:/Users/rosa/Documents/4CStudy/ClinicalData/investigation_dataset_4C_for_OLINK_WITH_SDS_Values_by_lms_method_for_cimt_and_pwv_(JH-16-NOV-2024).xlsx")

medication_file <- read_excel("C:/Users/rosa/Documents/4CStudy/ClinicalData/medication_dataset_4C_for_OLINK.xlsx")
patients_file <- read_excel("C:/Users/rosa/Documents/4CStudy/ClinicalData/patients_dataset_4C_for_OLINK.xlsx")

addedinfo_file <- read_excel("C:/Users/rosa/Documents/4CStudy/ClinicalData/4C_Serum_Olink.xlsx")
olink_data <- read.csv("Documents/4CStudy/Tables/OLINK_data.tsv",  sep = "\t",header=T) %>% arrange(pat_id) %>% rename(patid=pat_id) 

pat_with_olink <- olink_data$patid
```



## LVMI calculation 
based on paper xx 
```{r}
investigation_file <- investigation_file %>% mutate(lvmi_above95 = case_when(sex=="male" ~ 77.5265 - age_in_years_at_vdate
 * 15.8939 + (age_in_years_at_vdate^1.5) * 5.2322 - (age_in_years_at_vdate^2) * 0.4671 + 2.7380,
                             T ~  77.5265 - age_in_years_at_vdate * 15.8939 + (age_in_years_at_vdate^1.5) * 5.2322 - (age_in_years_at_vdate^2) * 0.4671) ) %>% mutate(lvmi_critical=lvmi >= lvmi_above95)


```

## eGFR check

Schaefer: Nine patients were excluded because of eGFR>60 ml/min per 1.73 m2, and seven were excluded due to missing eGFR at the first study visit, leaving a final cohort of 688 patients with CKD stages 3–5

comaprison with other 4C paper

```{r}
## 704 patients in line with schäfer
investigation_file %>% filter(month==0) %>% summarise(n())
baseline_file <- investigation_file %>% filter(month==0)

####### eGFR 
## distirbution of egfr values , 24 missing egfr values 
summary(baseline_file$egfr)
#baseline_file %>% filter(b_gfr > 60)
baseline_file %>% select(egfr,b_gfr)  ## -> similiar values, but more in egfr than b_gfr 

####### ckd stage 
table(baseline_file$ckd1)
nrow(baseline_file %>% filter(is.na(ckd1))) 

## 25 missing diagnosis , 16 more than in schäfer # schaefer numbers 3a: 50 	3b:216 	4:370 	5:52 	total 688

#baseline_file <- baseline_file  %>% filter(!is.na(ckd1))
baseline_file %>% filter(is.na(egfr))

```

what we could do is taken the b_egfr ( baseline egfr column) in case egfr is not given 
but then we have different numbers of ckd stages 
as followed: 

```{r}
## 14 with b_gfr  but not egfr value 
investigation_file %>% filter(month==0) %>% filter(is.na(egfr) & !is.na(b_gfr))
  
baseline_file <- investigation_file %>% filter(month==0) %>% mutate(egfr=case_when(is.na(egfr)~b_gfr, T~egfr)) %>% 
mutate(ckd1= case_when(is.na(ckd1) & egfr > 60 ~ "2 (60-<90)",
                         is.na(ckd1) & egfr < 60 & egfr > 45 ~ "3a (45-<60)",
                         is.na(ckd1) & egfr < 45 & egfr > 30 ~ "3b (30-<45)",
                         is.na(ckd1) & egfr < 30 & egfr > 15 ~ "4 (15-<30)",
                         is.na(ckd1) & egfr < 15 ~ "5 (<15)",
                         T ~ ckd1))  
baseline_file %>% filter(is.na(ckd1))
baseline_file  %>% pull(ckd1 ) %>% table(.)

## more 3b than in schaefer
```

### Added Info File

```{r}
table(addedinfo_file$Diagnose_original)
summary(addedinfo_file$egfr_original)

addedinfo_file %>% ggplot(aes(x=Diagnose_original, y=egfr))+ geom_boxplot()
baseline_file %>% ggplot(aes(x=ckd1, y=egfr)) + geom_boxplot()

baseline_file <- baseline_file %>% full_join(addedinfo_file, by="patid", suffix=c("", "_added")) 


summary(baseline_file$egfr - baseline_file$egfr_original)
baseline_file %>% select(ckd1,Diagnose_original,ckd_added,ckd_bedside,egfr,egfr_original,b_gfr,patid) %>% filter(is.na(ckd1)) 

## for the three cases of egfr and having a olink 
baseline_file <- baseline_file %>% mutate(egfr= case_when(is.na(ckd1)~ egfr_original, T~egfr)) %>% mutate(ckd1= case_when(is.na(ckd1) & egfr< 30 & egfr > 15 ~ "4 (15-<30)", T ~ckd1))

ckd1_numbers <- baseline_file %>% filter(patid %in% olink_data$pat_id) %>% count(ckd1)
ckd1_olink_numbers <- baseline_file %>% filter(patid %in% pat_with_olink) %>% count(ckd1)

```

## center site  & filter olink
only patients with olink sample (683)
adding country of origin as info 
```{r}
baseline_file %>% select(center,center_name) %>% distinct() %>% arrange(center)

baseline_file <- baseline_file  %>% filter(patid %in% pat_with_olink) %>% mutate(country= case_when( startsWith(center,"ank")|  startsWith(center,"is")| center %in% c("ada","izm","bu2","bur","gaz","iz2","mal","man","san") ~"Turkey", center %in% c("mue","ros","jen","lei","mar","ham","han","hei","col","erl","ess","fre","ber")~"Germany", center %in% c("bor","lyo","str")~"France", center %in% c("pad","mil","gen","rom","bol") ~"Italy", center %in% c("war","szc","cra","gda")~"Poland", center =="lon" ~"UK",center %in% c("inn","vie")~"Austria",center =="bel" ~"Serbia", T ~"others" )) 

baseline_file %>% pull(country) %>% table(.)
```

### country plot 

```{r}
country_colors <- c(
  Austria = "#FF4D6FFF",
  France = "#579EA4FF",
  Germany = "#DF7713FF",
  Italy = "#F9C000FF",
  others = "#86AD34FF",
  Poland = "#5D7298FF",
  Serbia = "#81B28DFF",
  Turkey = "#7E1A2FFF",
  UK = "#2D2651FF",
  notTurkey="#334e82" #"#644d8d"
)

# Prepare the data
plot_data <- baseline_file %>%
  distinct(patid, center, country) %>%
  mutate(center = gsub('[0-9]+', '', center)) %>% # Remove numbers from center names, merging centers from same city together
  count(country, center) %>%
  ## duplicate patids not from turkey, as "notTurkey"
  rbind(c("notTurkey", "", nrow(baseline_file %>% filter(country != "Turkey") %>% distinct(patid)))) %>%
  mutate(n = as.numeric(n)) %>%
  arrange(country, center) # Ensure centers are sorted within each country

# Generate gradient colors dynamically
# Create a list of gradients for each country
gradient_colors <- unlist(
  lapply(unique(plot_data$country), function(country) {
    base_color <- country_colors[country]
    centers <- plot_data %>% filter(country == !!country) %>% pull(center)
    
    # Create a gradient palette for the centers
    colors <- colorRampPalette(c(base_color, "white"))(length(centers)+1)
    
    # Assign center names to the colors
    setNames(colors, centers)
  })
)
# Map the gradient colors back to each center
plot_data <- plot_data %>%
  mutate(center_color = unlist(gradient_colors[match(country, names(gradient_colors))]))

# Create the bar plot
ggplot(plot_data, aes(x = reorder(country, n, sum), y = n, fill = center)) +
  geom_bar(position = "stack", stat = "identity") +
  coord_flip() +
  # Add labels for the center names on the stacked bars
  geom_text(
    aes(label = center),
    position = position_stack(vjust = 0.5), # Center labels within stacks
    size = 2,
    angle=90,
    color = "black"
  ) +
  # Use the dynamically generated gradient colors
  scale_fill_manual(values = gradient_colors) +
  labs(
    x = "",
    y = "Number of Patients",
  ) +
  theme_classic() + theme(legend.position = "none")

#ggsave("~/Documents/4CStudy/Plots/bar_country_distribution.pdf", width=6,height=3)

```


## 1 Descriptive Statistics 
all following descriptive statisticss for the patients with olink sample and having a eGFR at study entrance
### CKD and diagnose
distirbution of diagnoses in the different ckd stages

```{r}
patients_file  %>% right_join(.,  baseline_file %>% select(patid,ckd1,egfr), by="patid") %>% filter(patid %in% pat_with_olink & !is.na(ckd1)) %>%
  rename(postAKI_HUS=nephloss)  %>% mutate(diagnose2=case_when(diagnose2=="Nephloss"~"postAKI/HUS",T~diagnose2)) %>%
  mutate(diagnose2=factor(diagnose2, levels=rev(c("Other", "postAKI/HUS", "Glomerulopathies","Tubint", "CAKUT")))) %>%
  group_by(ckd1) %>% count(diagnose2)%>%  mutate(perc=n/sum(n) ) %>% arrange(ckd1,diagnose2) %>% 
  
  ggplot(aes(x=ckd1, y=perc, fill=diagnose2)) + 
  geom_col(position=position_stack()) + 
  geom_text(aes(label=n), position = position_stack(vjust =0.5), size =3.5, color="white") + 
  theme_classic()+ theme(axis.text.x = element_text(angle = 45,hjust=1)) + xlab("") + ylab("")+ 
  scale_fill_paletteer_d("PrettyCols::Bold", name="")

#ggsave("~/Documents/4CStudy/Plots/bar_CKD_diagnose.pdf", width=6,height=4)

```
eGFR values for the different diagnose groups 
```{r}
patients_file  %>% right_join(.,  baseline_file %>% select(patid,ckd1,egfr), by="patid") %>% filter(patid %in% pat_with_olink & !is.na(ckd1)) %>%
  rename(postAKI_HUS=nephloss)  %>% mutate(diagnose2=case_when(diagnose2=="Nephloss"~"postAKI/HUS",T~diagnose2)) %>%

  ggplot(aes(y=egfr, x =diagnose2, fill=diagnose2)) + 
  geom_boxplot()+
 theme_classic()+ theme(axis.text.x = element_text(angle = 45,hjust=1)) + xlab("") + ylab("")+ 
  scale_fill_paletteer_d("PrettyCols::Bold", guide="none") + ylab("eGFR")
  #scale_fill_paletteer_d("fishualize::Callanthias_australis",name="")
  #scale_fill_paletteer_d("nationalparkcolors::Badlands")

#ggsave("~/Documents/4CStudy/Plots/boxplot_egfr_diagnose.pdf", width=6,height=4)

```

### missing values 

```{r}
na_counts <- sapply(baseline_file, function(y) sum(length(which(is.na(y)))))
na_counts[c("lvmi", "pwv_sds_a", "imt_sds_a", "syssds", "diasds")]
```




### numeric variables

for all patients with eGFR value at study entry and olink sample

mean and sd values for
dist_ckd_diag_to_BL: distances in year between the ckd diagnosis and baseline assessment
age_in_years_at_vdate: age at baseline
gestage: gestational age 
bweight/height: birth weight and height 
egfr
blood values (salb,...)
blood pressure values (dia, sys)

```{r warning=FALSE}
numericvar <- c("heightsds","bmisds","age_in_years_at_vdate","egfr", "alburie", "urea", "salb", "bicar","hb","phosphaturie","diasds","syssds")

## mean for each ckd stage
patients_file  %>% right_join(.,  baseline_file %>% filter(patid %in% pat_with_olink &!is.na(ckd1)) %>% 
                                select(patid, ckd1, numericvar ), by="patid")  %>% mutate(dist_ckd_diag_to_BL=-age_in_years_at_ckd2date+age_in_years_at_vdate,egfr) %>% group_by(ckd1) %>% 
summarise(across(c(dist_ckd_diag_to_BL,numericvar), list(mean = mean, sd = sd),na.rm=T)) %>%
  left_join(BL_ckd1_olink_numbers,., by="ckd1")#%>% #write.table(.,  "~/Documents/4CStudy/Tables/cohortdescription_numericalvariables_n683_divided_CKDstages.tsv", row.names=F, quote=F,na="NA", sep="\t")


## mean all together
patients_file  %>% right_join(.,  investigation_file %>% filter(month==0) %>% select(patid, numericvar), by="patid") %>%   mutate(dist_ckd_diag_to_BL=-age_in_years_at_ckd2date+age_in_years_at_vdate,egfr) %>% 
summarise(across(c(dist_ckd_diag_to_BL,numericvar), list(mean = mean, sd = sd),na.rm=T))# %>% #write.table(.,  "~/Documents/4CStudy/Tables/cohortdescription_numericalvariables_n683.tsv", row.names=F, quote=F,na="NA", sep="\t")

```

## Define BL dataframe
```{r}
BL_data_bpoffice <- baseline_file %>%right_join(patients_file %>% select(-sex,-gender,-ethnicf,-ethnicm),., by="patid")

BL_data_bpoffice <- BL_data_bpoffice %>% mutate(CAKUT= case_when(diagnose2=="CAKUT"~"yes", T~"no"))
```

## 2 CVD endpoints
sds above 3 as "endpoint"
```{r}
BL_data_bpoffice <- BL_data_bpoffice %>% mutate(across(
    c(ends_with("sds_a"),syssds,diasds) , 
    ~ ifelse(. > 2, 1, 0),  # Check if the value is above 2, assign 1 if TRUE, 0 if FALSE
    .names = "{.col}_above2" # Create new columns with a suffix "_above2"
  )) %>% mutate(lvmi_critical= as.numeric(lvmi_critical))

```

## 3. Medication 

### Data Cleaning 
immune suppressants -> olink data 
antihypertensive medication for cvd endpoints (schaefer et al , preprint Doyon et al)

antihypertension treatment:
other diuretics, loop diuretics, alpha and beta blockers, calcium channel blockers, RAS inhibitors

other drugs:
laxatives, potassium-related medications, anti-epileptics

considering all mediaction, where the age at BL is between the start age and stop age of taken this medication
```{r}
## 351 patients taken medication when month==0, also when olink is taken 

# with 1 to 13 drugs, median 4 drugs 
summary(right_join(medication_file, investigation_file %>% filter(month==0 ) %>% select(patid,age), by="patid") %>% filter(age >=age_in_years_at_startdate & age <=age_in_years_at_stopdate) %>% count(patid) %>% pull(n))

## impact on olink and/or cvd measurements 
medication_at_BL<- right_join(medication_file, investigation_file %>% filter(month==0 ) %>% select(patid,age), by="patid") %>% filter(age >=age_in_years_at_startdate & age <=age_in_years_at_stopdate)
```

change the medigroup for some med ids
```{r}
medication_at_BL <- medication_at_BL %>% mutate(
      medigr= case_when(medid=="Bendroflumethiazide" ~  "Thiazide Diuretics",
      medid=="Polistiren sülfonat" | medid== "Sodium polystiren sulfonate" ~ "Ca-P-Bind",
      medid=="Choledtyramine" | medid=="Omega-3 fatty acids" ~ "Non-Statin Lipid Lowering Agent",
      medid=="Eculizumab" ~ "Immunsuppressiva",
      startsWith(medid,"Vitamin B") ~ "Vitamin B compounds",
      medid=="Shohl" ~ "Acidosis Treatment",
      endsWith(medid, "sone") ~ "Immunsuppressiva",
      is.na(medigr) ~ "others",
      T ~ medigr)) 

medication_at_BL <- medication_at_BL %>% mutate(medigr2= case_when(RAS=="yes"~ "RAS inhibitors", 
                                                                   endsWith(medigr,"alpha blocker")~ "Alpha-Blockers", 
                                                                   medigr=="Potassium sparing diuretics" | medigr=="Thiazide Diuretics"~"other diuretics",
                                                                   T~medigr))

medication_at_BL_wide <- medication_at_BL%>% select(patid, age_at_medication=age,medigr2) %>% distinct(patid, medigr2) %>% mutate(taken=1) %>%pivot_wider(names_from=medigr2, values_from=taken) %>% mutate_all( ~replace_na(.,0)) 

names(medication_at_BL_wide)[-1] <- paste0("medigr_", names(medication_at_BL_wide)[-1])

# add information to BL data frame
BL_data_bpoffice <- BL_data_bpoffice %>% left_join(medication_at_BL_wide, by="patid")
```


### Anti Hypertension Medications
define whether patient is on antihpyertensive medication 
```{r}
aht_drugs <- paste0("medigr_", c("Beta-Blocker", "RAS inhibitors","Loop Diuretics", "Alpha-Blockers", "Ca-Blocker","other diuretics"))

BL_data_bpoffice <- BL_data_bpoffice %>% rowwise()%>%  mutate(sum_medication = sum(c_across(starts_with("medigr")), na.rm = TRUE) ) %>%  mutate(sum_AHT = sum(c_across(aht_drugs), na.rm = TRUE) ) %>% mutate(AHT= as.numeric(sum_AHT>0))

```


### CVD and Hypertension Groups

hypertension classes 
four category : AHT treatment (yes/no) vs dia/sys > 2 
and binary class:( AHT treatment or dia>sys > 2 ) vs none hypertension signs
```{r}
BL_data_bpoffice <- BL_data_bpoffice %>% mutate(hyper_treat_group=  case_when(AHT==1 & (diasds >2 | syssds >2) ~ "treated_uncontrolled",
                                                    AHT==1 & (diasds <2 & syssds <2) ~ "treated_norm",
                                                    AHT!=1 & (diasds >2 | syssds >2) ~ "untreated_uncontrolled",
                                                    AHT!=1 & (diasds <2 & syssds <2) ~ "untreated_norm"
                                                  
                                                    ), hyper_class= case_when(AHT==1 | diasds >2 | syssds>2 ~ "yes", T~"no"))
```


## CVD manifestations
```{r}
BL_data_bpoffice <- BL_data_bpoffice %>% group_by(patid) %>% mutate(cvd_manifestation = sum(lvmi_critical, imt_sds_a_above2, pwv_sds_a_above2, na.rm=T) ) %>% ungroup()
```

## SAVE BL data bp office
used for all other analysis
```{r}
write.table( BL_data_bpoffice , "~/Documents/4CStudy/Tables/BL_data_bpoffice.tsv", row.names=F, quote=F,na="NA", sep="\t")
```


## Categorical Variables Statistics 

#### categorical values 

same procedure as for the numerical variables
overall patients and once separated by ckd stage 
calculating number and percentage of occurrence for each subcategory 
```{r}

categorical_columns <- c("sex", "diagnose2", "sga", "physact","country", "medigr_RAS.inhibitors", "hyper_class")

patients_file  %>% right_join( BL_data_bpoffice %>% filter(patid %in% olink_data$pat_id & !is.na(ckd1)) %>% select(patid, ckd1, physact,country, medigr_RAS.inhibitors, hyper_class) %>% mutate(across(c(medigr_RAS.inhibitors,hyper_class), as.character)) %>% filter(patid != pat_withoutegfr), by="patid") %>% group_by(ckd1)  %>%
pivot_longer(cols = all_of(categorical_columns), names_to = "variable", values_to = "category") %>%
  # Count occurrences of each category within each group and variable
  #filter(!is.na(category)) %>%
  mutate(total_patients = n_distinct(patid)) %>%
  count(ckd1, variable, category,total_patients) %>%
  # Calculate the percentage of each category within each group and variable
  mutate(percentage =( n / total_patients)* 100) %>%
  # Reshape data: make each category a separate column for each variable
  mutate(varcar=paste(variable,category,sep="_"))%>%
  select(-c(variable,category,total_patients)) %>%
  pivot_wider(
    names_from = c("varcar"),
    values_from = c(percentage,n),
  ) %>% select(sort(names(.))) %>% relocate(contains("sex"), .after=ckd1) %>%
  left_join(BL_ckd1_olink_numbers,., by="ckd1")  %>% write.table(.,  "~/Documents/4CStudy/Tables/cohortdescription_categoricalvariables_n683_divided_CKDstages.tsv", row.names=F, quote=F,na="NA", sep="\t")

patients_file  %>% right_join( BL_data_bpoffice%>% filter(patid %in% olink_data$pat_id & !is.na(ckd1))  %>% select(patid, ckd1, physact,country, medigr_RAS.inhibitors, hyper_class) %>% mutate(across(c(medigr_RAS.inhibitors,hyper_class), as.character)) %>% filter(patid != pat_withoutegfr), by="patid") %>%
pivot_longer(cols = all_of(categorical_columns), names_to = "variable", values_to = "category") %>%
  # Count occurrences of each category within each group and variable
  #filter(!is.na(category)) %>%
  mutate(total_patients = n_distinct(patid)) %>%
  count(variable, category,total_patients) %>%
  # Calculate the percentage of each category within each group and variable
  mutate(percentage =( n / total_patients)* 100) %>%
  # Reshape data: make each category a separate column for each variable
  mutate(varcar=paste(variable,category,sep="_"))%>%
  select(-c(variable,category,total_patients)) %>%
  select(sort(names(.))) %>% relocate(contains("sex"))  %>% write.table(.,  "~/Documents/4CStudy/Tables/cohortdescription_categoricalvariables_n683.tsv", row.names=F, quote=F,na="NA", sep="\t")

```



