---
title: "Cox Regression"
output:
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

includes the cox models with stratification for ckd and age and the HR plot for the CKD/UACR subgroups 


```{r libraries, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(readxl)
library(paletteer)
library(tibble)
library(mlr)
library(stringr)
library(survival)
library(purrr)
library(ggpubr)
library(survminer)
library(broom)
library(gridExtra)
library(viridisLite)
library(ggstats)
library(matrixStats)
library(ggrepel)
```

## 0 Functions
```{r include=FALSE}
extract_cox_results <- function(coxmodel){
  summary_cox <- summary(coxmodel)
hr <- exp(coef(coxmodel))  # HR = exp(beta)
ci <- exp(confint(coxmodel))  # 95% CI for HR
p_vals <- summary_cox$coefficients[, "Pr(>|z|)"]
p.adjusted <- p.adjust(p_vals, method = "BH")

# Create a data frame
results <- data.frame(
  Variable = names(hr),
  HR = hr,
  CI_lower = ci[, 1],
  CI_upper = ci[, 2],
  p_value = p_vals,
  p_adj=p.adjusted
)
results$p_label <- ifelse(results$p_value < 0.001, "<0.001", sprintf("%.3f", results$p_value))
results <-results %>% mutate( significant=(p_value <0.05), Variable=factor(Variable,levels=rev(rownames(results))))

return(results)
}



plot_HR_forest <- function(coxmodel,labels){

results <- extract_cox_results(coxmodel)

plot<- ggplot(results, aes(x = Variable, y = HR)) + 
  
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, color = "black") +  # Error bars for CI
   geom_point(size = 3, color = "blue", aes(alpha=significant)) + 
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +  # Reference line at HR = 1
  coord_flip() +  # Flip to horizontal
   geom_text(aes(label = paste0("p = ", p_label), x = Variable, y = max(CI_upper) + 0.15),  # Adjust position
            hjust = 0, size = 3, color = "black")+
  theme_classic() + scale_alpha_manual(values=c(0.4,1))+ 
  labs(x="" , y = "Hazard Ratio (HR)") + scale_x_discrete(labels=labels)+
  theme(axis.text = element_text(size = 10), legend.position = "none") 

return(plot)
}
```

## 1 Data

survival data: with survival time (time in years from study entrance
till endpoint is reached or study left) reachep: composite endpoint
including 50% eGFR loss, Transplantation or Dialysis

data was generated in SurvivalAnalysis_Preparation
```{r}
### defined inflammation score with a cutoff of 80th, based on egfr and proteinuria -> 660 patients 
coxdata_with_iscore <-  read.csv("~/Projects/4cstudy/Tables/survival_data.csv",  sep = ",",header=T)

## 5p , 80 cutoff
coxdata_with_iscore <- coxdata_with_iscore %>% mutate(i_score_5p_80= high_80_residual_pd_l1 + high_80_residual_cd40+high_80_residual_tnfrsf9+high_80_residual_cx3cl1+high_80_residual_il_15ra )

## 5p 50 cutoff
coxdata_with_iscore <- coxdata_with_iscore %>% mutate(i_score_5p_50=high_50_residual_pd_l1 + high_50_residual_cd40+high_50_residual_tnfrsf9+high_50_residual_cx3cl1+high_50_residual_il_15ra )
```
```{r}
## additional information about medication
BL_data_bpoffice <-read.csv2("~/Projects/4cstudy/Tables/BL_data_bpoffice.csv",  sep = ",",header=T) %>% filter(patid != 737) %>% mutate(across(everything(), ~ type.convert(.x, as.is = TRUE)))
```


## 2 Cox Regression

survival time: time from study entrance till endpoint is reached

endpoint (reachep): 50% eGFR loss, dialysis or transplantation

goal: create a good predictive cox model

including inflammation score accounting for other factors and predictors
of CKD progression

-\> quite a lot of variables, proportional Hazard assumption?

### 2.1 Straight-forward Model

adding all relevant variables (based on literature and results of Random
Survival Forest variable importance):

-   eGFR

-   proteinuria

-   age at study entrance (BLage)

-   underlying diseases (CAKUT,Glomerulonephritis,..), diagnose2

-   sex

-   hemoglobin (hb)

-   blood phosphorus (phos)

-   serum albumin (salb)

-   our defined inflammation score : i-score

```{r}
formula <- as.formula("Surv(survival_time,reachep)~ egfr +log(alburie+0.01)+ BLage+ diagnose2+ sex  + log(salb)+diasds+ +i_score")

coxmodel_simple <-  coxph(  formula=formula, 
                              
                          data= coxdata_with_iscore)
cox.zph(coxmodel_simple)
plot(cox.zph(coxmodel_simple))
```

in particular beta coefficient of eGFR seems to increase with time

age fluctuates and alburie has as well a weird shape

-\> trying strata to control for different effect of eGFR, proteinuria
and age

### 2.2 Stratified Model

-   stratifying based on CKD stage (2,3a,3b,4,5), ckd1

-   based on age: younger than 12 (median) or older than 12

-   and strata for proteinuria in (A1: \<3),(A2: \>3 & \< 30) ,(A3:
    \>30)

-   -\> different baseline hazard function depending on kidney function
    at study entrance (eGFR and proteinuria) and age

#### Strata Investigation

```{r}
coxdata_with_iscore<-coxdata_with_iscore %>% mutate( age2= if_else(BLage <= 12.205,"young","old"), 
                                                     acr_grp= case_when(alburie< 30 ~"A1", alburie>300~"A3",T~"A2") )

ckd2stage_pats <- coxdata_with_iscore %>% filter(egfr>60) %>% pull(patid)
```

risk of overfitting?

as strata of the three variables: ckd stage, proteinuria and age results
in a lot of small strata

```{r}
coxdata_with_iscore %>% mutate(strata=paste(age2,ckd1,acr_grp,sep=",")) %>% dplyr::count(strata) %>% arrange(n)
```

not stratifying by proteinuria?

```{r}
coxdata_with_iscore %>% mutate(strata=paste(age2,ckd1,sep=",")) %>% dplyr::count(strata) %>% arrange(n)

```

kaplan meier curves for the strata

```{r}
ggsurvplot(surv_fit( as.formula("Surv( survival_time,reachep)~ strata"), 
                     data= coxdata_with_iscore %>% filter(!patid %in% ckd2stage_pats) %>% mutate(strata=paste(age2,ckd1,sep=","))), 
           risk.table=T, ggtheme=theme_classic(), palette = c( "#FCAADEFF","#EF5FAFFF","#C31E6EFF","#7F0038FF", "#99CE64FF", "#659A32FF", "#326812FF", "#033800FF"))
```

clearly different curves for the ckd stages and also in some ckd stages
differences in age, in particular stage 4

#### final strata model

fitting it without ckd stage 2 , not enough patients (n=8) for having
their own strata

```{r}
formula <- as.formula("Surv(survival_time,reachep)~egfr +BLage+ sex+ log(alburie+0.01)+ diagnose2+log(salb)+ phos+diasds+i_score+ survival::strata(ckd1,age2)")

coxmodel_withstrata <-  coxph(  formula=formula, 
                              
                          data= coxdata_with_iscore %>% filter(!patid %in% ckd2stage_pats))


cox.zph(coxmodel_withstrata)
plot(cox.zph(coxmodel_withstrata))
```

improved schonefeld plots for all variables

age looks quite flat now and also improvement in eGFR and other variables


```{r warning=FALSE}
AIC(coxmodel_simple,coxmodel_withstrata)
```

```{r}
names(coxmodel_withstrata$coefficients)
labels <- c("egfr"="eGFR","BLage"="age at BL", "sexmale"="sex", "log(alburie + 0.01)"="log(alburie + 0.01)",
            "diagnose2Glomerulopathies"="glomerulopathies","diagnose2Nephloss"="postAKI/HUS",
            "diagnose2Other"="other diagnose", "diagnose2Tubint"="Tubint","log(salb)"="log(salb)", "hb"="hemoglobin","phos"="phosphorus", "diasds"="dia. SDS", "i_score"= "i-score")

plot_HR_forest(coxmodel_withstrata,labels=labels) + ylim(0.12,3.6)
```
#### additive approach
```{r}
########## only i-score
formula <- as.formula("Surv(survival_time,reachep)~i_score+ survival::strata(ckd1,age2)")

coxmodel_only_iscore <-  extract_cox_results( coxph(  formula=formula, data= coxdata_with_iscore %>% filter(!patid %in% ckd2stage_pats))) %>% mutate(model="only i-score")

######### i-score + KFRE
formula <- as.formula("Surv(survival_time,reachep)~i_score+ egfr +BLage+ sex+ log(alburie+0.01)+ survival::strata(ckd1,age2)")

coxmodel_iscore_KFRE <-   extract_cox_results(coxph(  formula=formula, data= coxdata_with_iscore %>% filter(!patid %in% ckd2stage_pats))) %>% mutate(model="KFRE")

######### i-score + phos + salb
formula <- as.formula("Surv(survival_time,reachep)~i_score+ +log(salb)+ phos + survival::strata(ckd1,age2)")

coxmodel_iscore_blood <-  extract_cox_results( coxph(  formula=formula, data= coxdata_with_iscore %>% filter(!patid %in% ckd2stage_pats)))%>%
  mutate(model="blood")


######### i-score + all + RAS
formula <- as.formula("Surv(survival_time,reachep)~egfr +BLage+ sex+ log(alburie+0.01)+  diagnose2+log(salb)+  phos+diasds+ RAS+i_score+ survival::strata(ckd1,age2)")

coxmodel_iscore_RAS <-  extract_cox_results( coxph(  formula=formula, data= coxdata_with_iscore %>% 
                                                       left_join(BL_data_bpoffice %>% select(patid,RAS=medigr_RAS.inhibitors)) %>% 
                                                       filter(!patid %in% ckd2stage_pats))) %>% mutate(model="RAS + all")

allresults <- rbind(coxmodel_iscore_RAS,coxmodel_iscore_blood,coxmodel_iscore_KFRE ,coxmodel_only_iscore,
                    extract_cox_results(coxmodel_withstrata) %>% mutate(model="all"))%>% mutate(model=factor(model, levels = rev(c("only i-score","KFRE","blood","all", "RAS + all"))))

```

```{r}
p<- allresults %>% filter(Variable=="i_score" & model!="RAS + all") %>% 
  ggplot(aes(x=model,y=HR)) + 
    
geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, color = "black") +  # Error bars for CI
   geom_point(size = 3, color = "blue", aes(alpha=significant)) + 
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +  # Reference line at HR = 1
  coord_flip() +  # Flip to horizontal
   geom_text(aes(label = paste0("p = ", p_label), x = model, y = max(CI_upper) + 0.02),  # Adjust position
          hjust = 0, size = 3, color = "black")+
  theme_classic() + scale_alpha_manual(values=c("FALSE"=0.4,"TRUE"=1))+ 
  labs(x="" , y = "Hazard Ratio (HR)") + scale_x_discrete(labels=labels)+
  theme(axis.text = element_text(size = 10), legend.position = "none") 

p_supplement <- allresults %>% filter(Variable=="i_score" ) %>%
  ggplot(aes(x=model,y=HR)) + 
geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, color = "black") +  # Error bars for CI
   geom_point(size = 3, color = "blue", aes(alpha=significant)) + 
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +  # Reference line at HR = 1
  coord_flip() +  # Flip to horizontal
   geom_text(aes(label = paste0("p = ", p_label), x = model, y = max(CI_upper) + 0.02),  # Adjust position
          hjust = 0, size = 3, color = "black")+
  theme_classic() + scale_alpha_manual(values=c("FALSE"=0.4,"TRUE"=1))+ 
  labs(x="" , y = "Hazard Ratio (HR)") + scale_x_discrete(labels=labels)+
  theme(axis.text = element_text(size = 10), legend.position = "none") 

```




### other i-scores

#####  4p , cut off 80, category
```{r}
## 4p , cut off 80, categrocial
formula_i80_cat<- as.formula("Surv(survival_time,reachep)~egfr +BLage+ sex+ log(alburie+0.01)+  diagnose2+log(salb)+diasds+ +hb+phos+i_score_80_cat + survival::strata(ckd1,age2)")

coxmodel_withstrata_i80_cat <-  coxph(  formula=formula_i80_cat, 
                              
                          data= coxdata_with_iscore %>% mutate(i_score_80_cat=case_when(i_score_80==0~"0", i_score_80>0 & i_score_80<=1 ~"1", T~"2-4")) %>% filter(!patid %in% ckd2stage_pats))

plot_HR_forest(coxmodel_withstrata_i80_cat,labels=labels) + ylim(0.2,3.6)
```


```{r}
## 4p , pseudo continous
formula_i80_cat<- as.formula("Surv(survival_time,reachep)~egfr +BLage+ sex+ log(alburie+0.01)+  diagnose2+log(salb)+diasds+ +hb+phos+i_score_80_cat + survival::strata(ckd1,age2)")

coxmodel_withstrata_i80_cat <-  coxph(  formula=formula_i80_cat, 
                              
                          data= coxdata_with_iscore %>% mutate(i_score_80_cat=case_when(i_score_80==0~0, i_score_80>0 & i_score_80<=2 ~1, T~2)) %>% filter(!patid %in% ckd2stage_pats))

plot_HR_forest(coxmodel_withstrata_i80_cat,labels=labels) + ylim(0.2,3.6)
```

#####  4p , cut off 50
```{r}
## 4p , cut off 50
formula_i50 <- as.formula("Surv(survival_time,reachep)~egfr +BLage+ sex+ log(alburie+0.01)+  diagnose2+log(salb)+diasds+ +hb+phos+i_score_50 + survival::strata(ckd1,age2)")

coxmodel_withstrata_i50 <-  coxph(  formula=formula_i50, 
                              
                          data= coxdata_with_iscore %>% filter(!patid %in% ckd2stage_pats))
plot_HR_forest(coxmodel_withstrata_i50,labels=labels) + ylim(0.2,3.6)

```

##### 5p , cut off 80
```{r}
## 5p , cut off 80
formula_5p_i80 <- as.formula("Surv(survival_time,reachep)~egfr +BLage+ sex+ log(alburie+0.01)+  diagnose2+log(salb)+diasds+ +hb+phos+i_score_5p_80 + survival::strata(ckd1,age2)")

coxmodel_withstrata_5p_i80 <-  coxph(  formula=formula_5p_i80, 
                              
                          data= coxdata_with_iscore %>% filter(!patid %in% ckd2stage_pats))
plot_HR_forest(coxmodel_withstrata_5p_i80,labels=labels) + ylim(0.2,3.6)
```

## 3 Plot Stratification
visualizing hazard ratio of i-score for each ckd stage and albuminuria group
solely plotting, for some ckd stage not enough sample size, too uncertain, great range of CIs

```{r}
## simple cox model for each subset (ckd and acr)
get_hazardratio <- function(df,iscore="i_score_50"){

form_text <- paste0("Surv( survival_time,reachep)~  BLage +sex + egfr + log(alburie +0.01) + strata(age2)+", iscore)
  
cox_model<- coxph(as.formula(form_text), data=df )
cox_sum<- summary(cox_model)

hr <- coef(cox_sum)[iscore, "exp(coef)"]
p_value <-cox_sum$coefficients[iscore, "Pr(>|z|)"]
ci_low <- exp(confint(cox_model))[iscore, 1]
ci_up<- exp(confint(cox_model))[iscore, 2]
size<- cox_model[["n"]]

return(data.frame( hr=hr,p_val=p_value,CI_low=ci_low,CI_up=ci_up,n=size))
}
```

```{r}
allstages_cueniform_hr <- list()

for (stage in rev(unique(coxdata_with_iscore$ckd1))){
  for(alb_grp in rev(unique(coxdata_with_iscore$acr_grp))){
    
    
    data<- coxdata_with_iscore %>% filter(ckd1==stage & acr_grp==alb_grp)
  
    if(nrow(data)>5){
    hrs <- get_hazardratio(df=data ,iscore="i_score_80") %>%  
      mutate(ckdstage=stage,alb_grp=alb_grp)
    }
    else{
     hrs<- data.frame(hr=NA,p_val=NA,CI_low=NA,CI_up=NA,n=nrow(data),ckdstage=stage,alb_grp=alb_grp)
    }
  

  allstages_cueniform_hr[[paste0(stage,":",alb_grp )]]<- hrs
  
  
  }}

hr_cueniform_data<- bind_rows(allstages_cueniform_hr)


hr_cueniform_data %>% filter(n>10)
```

```{r}
plot_cueninform <- function(data){
maxhr <- max(data %>% filter(!is.infinite(CI_up)&CI_up <100000 & !is.na(hr)) %>% pull(hr))
minhr <- min(data %>% filter( !is.infinite(CI_up)& CI_up <100000 & !is.na(hr)) %>% pull(hr))

print(minhr)
if (minhr > 1){
  cols <- c("#BB292CFF", "#FFD353FF")
  vals<-  scales::rescale(c(maxhr+0.3, minhr),
                        limits = c(minhr,maxhr+0.3))
  brks<- c(minhr,(minhr+ maxhr)/2,maxhr, maxhr + 0.3)
  lbls<-c(sprintf("%.1f", minhr), sprintf("%.1f", (minhr+ maxhr)/2), sprintf("%.1f", maxhr), paste0("> ",round(maxhr,1)))
}
else{
  cols<- c("#BB292CFF", "#FFD353FF", "#72874EFF", "#466532FF")
  vals <-  scales::rescale(c(maxhr+0.3, 1,0.99, minhr),
                        limits = c(minhr,maxhr+0.1))
  brks<- c(minhr,1,(1+ maxhr)/2, maxhr,maxhr+0.3)
  lbls<- c(sprintf("%.1f", minhr), "1",sprintf("%.1f", (1+ maxhr)/2), sprintf("%.1f", maxhr), paste0("> ",round(maxhr,1)))
   # lbls<- c(sprintf("%.1f", minhr),"1",sprintf("%.1f", (1+ maxhr)/2), sprintf("%.1f", maxhr))
}

### breaks for sample size
sizebreaks <- c(min(data %>% pull(n)),median(data %>% pull(n)),100,max(data %>% pull(n)))

plot<-  ggplot(data %>% filter( !is.na(hr)) %>% mutate( direction=if_else(hr>1,"HR>1","HR<1"), hr= if_else(is.infinite(CI_up)| CI_up >100000,maxhr+0.3,hr), ckdstage= factor(ckdstage, levels=unique(coxdata_with_iscore$ckd1))), aes(x = alb_grp, y = ckdstage)) +
  geom_point(aes(
    color = hr ,
    fill=hr,
    shape = direction,  ),size=4 ,stroke = 0.9) +
  geom_text_repel(aes(label=n),min.segment.length="",vjust=2.5,size=3)+
  scale_shape_manual(values = c("HR<1"=25,"HR>1"=24),name="") +  # 25 (downward triangle), 24 (upward triangle)
   scale_fill_gradientn(colors = cols, 
                        values =vals,
                        breaks = brks,  # Custom ticks
                        labels = lbls # Custom labels
  ) + 
    scale_color_gradientn(colors = cols, 
                        values = vals) + 
  
 scale_x_discrete(position = "top")+
  theme_minimal() + scale_size_continuous(breaks = sizebreaks)+
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()  )+
  labs(
    title = "",
    x = "Albuminuria Group",
    y = "CKD stage",
    fill = "Magnitude of HR",
    shape= "HR >1",
    size = "Sample Size"
  ) + geom_stripped_rows() +guides(colour = "none")
 

 return(plot)
}
```


```{r}
plot_cueninform(hr_cueniform_data %>% filter(n>10 )) 
ggsave(paste0("~/Projects/4cstudy/Plots/SurvivalAnalysis/Figure4/HRcueniform_stratification_4p_i80_number_samplesize.pdf"),plot_cueninform(hr_cueniform_data %>% filter(n>10 )), width=6.8,height=4.9)
```
