---
title: "Linear Models & Contrasts"
output: html_notebook
---

This script contains the analysis and plot of figure 2:

linear models of eGFR/UACR for the different OLINK proteins as well as contrast test for investigating the influence of patient's origin (country) and diseases

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(tibble)
library(metadeconfoundR)
library(ggpubr)
library(readxl)
library(paletteer)
library(stringr)
library(glmnet)
library(mlr)
theme_classic_modified <- function(){
  theme_classic() %+replace%
    theme(strip.placement="outside", 
                           strip.text.y.left = element_text(angle = 0,size=11),
                           strip.background.y=element_rect(color = NA), 
                           strip.background.x=element_rect(color = NA),
                           legend.position = "none") 
}
```
analysis showing in figure 2 
associations between eGFR and other clinical variables at study entrance to olink protein profile
## Data Input
```{r message=FALSE, warning=FALSE}
pat_withoutegfr <- xxx

### baseline table: contains all clinical variables at month 0 for each patient 
BL_data_bpoffice <-read.csv2("~/Projects/4cstudy/Tables/BL_data_bpoffice.csv",  sep = ",",header=T) %>% filter(patid != pat_withoutegfr) %>% mutate(across(everything(), ~ type.convert(.x, as.is = TRUE)))

### investigation file from ckd progression, clinical data from all visits month 0 -120 
investigation_file <-read.csv("~/Projects/4cstudy/Tables/investigation_file.csv",  sep = ",",header=T) %>% filter(!is.na(month) & month!="FALSE")%>% mutate(across(everything(), ~ type.convert(.x, as.is = TRUE))) 

### from ckd progression 
# contains the first month where the patient reached one of the endpoints: ckd50%, tx, dialysis
reached_endpoint <- reached_endpoint <- read.csv("~/Projects/4cstudy/Tables/reached_endpoint.csv", sep = ",", header=T) %>% mutate(across(everything(), ~ type.convert(.x, as.is = TRUE)))


### olink data

olink_data <- read.csv("~/Projects/4cstudy/Tables/OLINK_data.csv",  sep = ",",header=T) %>% rename(patid=pat_id) %>%arrange(patid) %>% mutate(across(everything(), ~ type.convert(.x, as.is = TRUE))) %>% filter(patid != 737)

### labels 
labels <- read_excel("~/Projects/4cstudy/ClinicalData/analyte_abbrev_HB.xlsx") 
```

## Beta Coefficient Analysis

### Function 
running linear model with eGFR as main predictor, other predictors as adjustment possible
scaling analyte yes or no 
returns beta estimate and CI for eGFR as well as wald test p-value, and model fit (r-square)

data: olink protein data frame
predictors: data frame of explaintory variables of interest (egfr,diagnose,...)
start, end : index of protein columns in the data frame 
```{r}
run_coefficent_egfr<- function(data, predictors, start,end, scale=T,mainpred="egfr"){
  res <- list()
  j<-1
  ## build linear model for each protein 
  for( i in start:end){
    
    y <- data.matrix(data[,i])
    analyte<- names(data)[i]
    ## build model 
    if(scale){
       lm=(lm(scale(analyte)~.,predictors %>% mutate(analyte=y)))
    }
    else{
       lm=(lm(analyte~.,predictors %>% mutate(analyte=y)))
    }
   
    fit = summary(lm)
    
    ## get confidence interval for eGFR coefficent
    confint_egfr <- data.frame(confint(lm) )%>% tibble::rownames_to_column("predictor")
   
    confint_egfr <- confint_egfr%>% filter(predictor==!!(mainpred))
    
    # save results of lm 
    df<- data.frame(analyte=analyte,
                    adjRsqr=fit$adj.r.squared,
                    p_valRsqr=pf(fit$fstatistic[1],fit$fstatistic[2],fit$fstatistic[3],lower.tail = F), pval=fit$coefficients[mainpred,4] ,
                    estimate=fit$coefficients[mainpred,1], 
                    down=confint_egfr[,2],up=confint_egfr[,3])

    res[[j]]<- df
    j <- j+1
  }
  
  res <- bind_rows(res)
  ## pval over all protein lms
  res$FDR <-p.adjust(as.numeric(res$pval), method="fdr")
  return(res)
}
```

### Build eGFR Models

three models:
- only eGFR 
- eGFR + "demographic variables": sex + age + bmisds + country
- eGFR + diagnose

for all models scaled npx-values for the proteins (making the beta coefficietnes more comparable)

model 1: only egfr
```{r}
### only egfr
  only_scaled_egfr_sa_lm_res <- run_coefficent_egfr(data = olink_data %>% filter(patid != pat_withoutegfr) %>% arrange(patid),  
                                                    predictors =  BL_data_bpoffice %>%arrange(patid) %>% dplyr::select(egfr) , start=82,end=158) 

only_scaled_egfr_sa_lm_res <-only_scaled_egfr_sa_lm_res %>%mutate( analyte=str_replace(analyte,"^rawscale_","")) %>% mutate(across(c(estimate,up,down), as.numeric)) 
```

model 2: egfr + demo
adding predictors: sex, age, bmi, country 
always scaling analytes

transforming country and sex to dummy variables 
```{r}
### egfr + demographs + country  
demo_egfr_sa_lm_res <- run_coefficent_egfr(data = olink_data %>% arrange(patid), 
                                  predictors =  BL_data_bpoffice   %>% dplyr::select(egfr,age,bmisds,sex, starts_with("country")) %>% fastDummies::dummy_columns(. ,select_columns = c("country", "sex"),remove_selected_columns = T) %>% select(-sex_female), 82,158) %>%mutate( analyte=str_replace(analyte,"^rawscale_","")) 

demo_egfr_sa_lm_res <- demo_egfr_sa_lm_res %>% mutate(across(c(estimate,up,down), as.numeric))
```

model 3: egfr + diagnose 

adding diagnose to the model , also here transforming to mulitiple dummy variables
```{r}
### egfr +diagnose 
diagnose_egfr_sa_lm_res <- run_coefficent_egfr(data = olink_data %>% arrange(patid), 
                                  predictors =  BL_data_bpoffice   %>% dplyr::select(egfr, diagnose2)  %>% fastDummies::dummy_columns(. ,select_columns = c("diagnose2"),remove_selected_columns = T) , 82,158) %>%mutate( analyte=str_replace(analyte,"^rawscale_","")) 


diagnose_egfr_sa_lm_res <- diagnose_egfr_sa_lm_res %>% mutate(across(c(estimate,up,down), as.numeric))
```
model 4: alburie

```{r}
### only alburie
  only_alburie_sa_lm_res <- run_coefficent_egfr(data = olink_data %>% filter(patid != pat_withoutegfr) %>% arrange(patid),  
                                                    predictors =  BL_data_bpoffice %>%arrange(patid) %>% mutate(alburie=-log(alburie+0.01)) %>%dplyr::select(alburie),mainpred = "alburie" , start=82,end=158) 

only_alburie_sa_lm_res <-  only_alburie_sa_lm_res %>%mutate( analyte=str_replace(analyte,"^rawscale_","")) %>% mutate(across(c(estimate,up,down), as.numeric)) 
```

model 5: egfr + alburie

```{r}
### egfr+ alburie
  egfr_alburie_sa_lm_res <- run_coefficent_egfr(data = olink_data %>% filter(patid != pat_withoutegfr) %>% arrange(patid),  
                                                    predictors =  BL_data_bpoffice %>%arrange(patid) %>% mutate(alburie=-log(alburie+0.01)) %>%dplyr::select(alburie,egfr), start=82,end=158) 

egfr_alburie_sa_lm_res <-  egfr_alburie_sa_lm_res %>%mutate( analyte=str_replace(analyte,"^rawscale_","")) %>% mutate(across(c(estimate,up,down), as.numeric)) 

## estimates for alburie
  alburie_egfr_sa_lm_res <- run_coefficent_egfr(data = olink_data %>% filter(patid != pat_withoutegfr) %>% arrange(patid),  
                                                    predictors =  BL_data_bpoffice %>%arrange(patid) %>% mutate(alburie=-log(alburie+0.01)) %>%dplyr::select(alburie,egfr), start=82,end=158, mainpred = "alburie") 

  alburie_egfr_sa_lm_res <-    alburie_egfr_sa_lm_res %>%mutate( analyte=str_replace(analyte,"^rawscale_","")) %>% mutate(across(c(estimate,up,down), as.numeric)) 
```

model 6: all 

```{r}
### egfr+ alburie
  egfr_all_sa_lm_res <- run_coefficent_egfr(data = olink_data %>% filter(patid != pat_withoutegfr) %>% arrange(patid),   predictors =  BL_data_bpoffice   %>% dplyr::select(egfr,alburie,age,bmisds,sex,diagnose2, starts_with("country")) %>% fastDummies::dummy_columns(. ,select_columns = c("country","diagnose2", "sex"),remove_selected_columns = T) %>% mutate(alburie=-log(alburie+0.01)) %>% select(-sex_female), start=82,end=158) 

  egfr_all_sa_lm_res <-   egfr_all_sa_lm_res %>%mutate( analyte=str_replace(analyte,"^rawscale_","")) %>% mutate(across(c(estimate,up,down), as.numeric)) 

## estimates for alburie
  alburie_all_sa_lm_res <- run_coefficent_egfr(data = olink_data %>% filter(patid != pat_withoutegfr) %>% arrange(patid),   predictors =  BL_data_bpoffice   %>% dplyr::select(egfr,alburie,age,bmisds,sex, starts_with("country"),diagnose2) %>% fastDummies::dummy_columns(. ,select_columns = c("country", "sex","diagnose2"),remove_selected_columns = T) %>% mutate(alburie=-log(alburie+0.01)) %>% select(-sex_female), start=82,end=158,mainpred = "alburie") 
  
  alburie_all_sa_lm_res <-     alburie_all_sa_lm_res %>%mutate( analyte=str_replace(analyte,"^rawscale_","")) %>% mutate(across(c(estimate,up,down), as.numeric)) 
```

### Visualize Beta Coef Plot
figure 2 plot

pretty names for the proteins
```{r}
labels <- setNames( labels$name,labels$analyte)
```

#### egfr + UACR coef
```{r}
## bind results of all three models
allresults <-  
  ## all model, coef for egfr
  egfr_all_sa_lm_res  %>% mutate(model="full",var="eGFR")   %>% 
   ## all model, coef for alburie
   rbind(.,alburie_all_sa_lm_res %>% mutate(model="full", var="UACR")) %>%
  ## only egfr
  rbind(.,only_scaled_egfr_sa_lm_res %>% mutate(model="eGFR", var="eGFR")) %>%
  ## only alburie
  rbind(.,only_alburie_sa_lm_res %>% mutate(model="UACR", var="UACR")) %>%
  ## egfr + alburie: egfr beta coef
   rbind(.,egfr_alburie_sa_lm_res %>% mutate(model="eGFR + UACR",var="eGFR")) %>%
  rbind(.,alburie_egfr_sa_lm_res %>% mutate(model="eGFR + UACR",var="UACR")) %>% 
  ## filter for FDR 
  group_by(analyte)%>%  filter(any(FDR < 0.05)) %>% mutate(significant=FDR<0.05) %>%
  ## naming models for plot
  mutate(model=factor(model, level=c( "eGFR","UACR","eGFR + UACR", "full"))) 

analyte_estimate_size <- allresults %>% filter(var=="eGFR") %>% group_by(analyte) %>% slice_max(abs(estimate)) %>% ungroup()%>% arrange(-estimate) %>% pull(analyte)
```

#### figure 2 plot
```{r}
 allresults %>% mutate(analyte=factor(analyte, level=analyte_estimate_size))%>%
  ## plot lolliplots/forest
ggplot(aes( y=analyte)) +
  geom_linerange(aes(xmin=down, xmax=up,color=var,alpha=significant) ) + 
  geom_point(aes(x=estimate,color=var,alpha=significant), shape=16, size=3) +
  facet_grid(.~ model, scales = "free")+
  ## aesethic: 
  theme_classic() + theme(legend.position = "top") + 
  ggstats::geom_stripped_rows()  + 
  scale_color_manual(values=c( "eGFR"="#A8CDECFF", "UACR"="#F6955EFF"), name="") + 
  scale_y_discrete(labels= labels) +
  scale_alpha_manual(values=c(0.4,1), labels=c("FDR< 0.05","FDR > 0.05" )) +
  labs(y=NULL ,
       color="coef",
       alpha=NULL,
       x=" beta coefficients of eGFR and UACR")

ggsave("~/Projects/4cstudy/Plots/Figure2/forest_betacoef_eGFR_UACR_alone_combined_full_sortedegfr.pdf", width=8,height=7.5)
```

####  alburie based
```{r}
analyte_estimate_size <- only_alburie_sa_lm_res  %>% arrange(estimate) %>% pull(analyte)
alburie_egfr_sa_lm_res  %>% mutate(model="alburie +egfr",var="alburie")   %>%  rbind(.,only_alburie_sa_lm_res %>% mutate(model="alburie",var="alburie")) %>%
 rbind(.,alburie_all_sa_lm_res%>% mutate(model="full model",var="alburie") )%>%
  ## filter for FDR 
  group_by(analyte)%>%  filter(any(FDR < 0.05)) %>% mutate(significant=FDR<0.05) %>% ## naming models for plot
  mutate(analyte=factor(analyte,levels = analyte_estimate_size)) %>%
  ## plot lolliplots/forest
ggplot(aes( y=analyte)) +
  geom_point(aes(x=estimate,color=significant), shape=16, size=3) +
  geom_linerange(aes(xmin=down, xmax=up,color=significant) ) + 
  facet_grid(.~ model , scales = "free_x")+
  ## aesethic: 
  theme_classic_modified() + theme(legend.position = "top") + ylab("") + xlab(" beta coefficients of log(albuminuria)") + 
  ggstats::geom_stripped_rows()  + 
  scale_color_manual(values=c( "grey","black"), labels=c("non-signif.", "signif."), name="") +  scale_y_discrete(labels= labels)
```
```{r}
analyte_estimate_size <- only_scaled_egfr_sa_lm_res  %>% arrange(-estimate) %>% pull(analyte)
egfr_alburie_sa_lm_res  %>% mutate(model="egfr + alburie",var="alburie")   %>%  rbind(.,only_scaled_egfr_sa_lm_res %>% mutate(model="egfr",var="alburie")) %>%
 rbind(.,egfr_all_sa_lm_res%>% mutate(model="full model",var="alburie") )%>%
  ## filter for FDR 
  group_by(analyte)%>%  filter(any(FDR < 0.05)) %>% mutate(significant=FDR<0.05) %>% ## naming models for plot
  mutate(analyte=factor(analyte,levels = analyte_estimate_size)) %>%
  ## plot lolliplots/forest
ggplot(aes( y=analyte)) +
  geom_point(aes(x=estimate,color=significant), shape=16, size=3) +
  geom_linerange(aes(xmin=down, xmax=up,color=significant) ) + 
  facet_grid(.~ model , scales = "free_x")+
  ## aesethic: 
  theme_classic_modified() + theme(legend.position = "top") + ylab("") + xlab(" beta coefficients of eGFR") + 
  ggstats::geom_stripped_rows()  + 
  scale_color_manual(values=c( "grey","black"), labels=c("non-signif.", "signif."), name="") +  scale_y_discrete(labels= labels)
```



### Splitted in Diagnose 

run the egfr model for each diagnose group separately 
comparing egfr beta estimated between the disease groups 
```{r}
## dataframe to save all results for each diagnose and protein combination
diagnose_linearmodel_splitted <-  data.frame(matrix(ncol=8,nrow=0))
names(diagnose_linearmodel_splitted) <-   c("analyte", "adjRsqr", "p_valRsqr", "pval", "estimate", "down", "up", "diagnose")

# iterate, run model for each disease group
for (dia in  levels(factor(BL_data_bpoffice$diagnose2)) ){
  ## get subset from clinical data 
  BL_split <- BL_data_bpoffice %>% filter(diagnose2==dia) %>% arrange(patid)
  ## run egfr model protein ~ eGFR
 res<-  run_coefficent_egfr(data = olink_data %>% filter(patid %in% BL_split$patid) %>% arrange(patid), 
                                  predictors =  BL_split  %>% dplyr::select(egfr) , 82,158) 
 
 res <- res %>%mutate( analyte=str_replace(analyte,"^rawscale_","")) %>% mutate(across(c(estimate,up,down), as.numeric)) %>% mutate(diagnose=dia)

# save results
 diagnose_linearmodel_splitted <- diagnose_linearmodel_splitted %>% rbind(res)
  
}
```

plot results as forest plot for all FDR survivors

same plot, once all in one "column", one plot
```{r}
#filter for signifiance
diagnose_linearmodel_splitted %>%  group_by(analyte)%>% filter(any(FDR < 0.05)) %>% mutate(significant=FDR<0.05) %>% 
# plot , separation in groups by setting color=diagnose
ggplot(aes( y=reorder(analyte, -as.numeric(estimate)))) +
  geom_point(aes(x=estimate,color=diagnose, alpha=significant),position = position_dodge(width = 1), shape=16, size=3) +
  geom_linerange(aes(xmin=down, xmax=up,color=diagnose, alpha=significant),position = position_dodge(width = 1) ) +
  # aesthetics
  theme_classic() + ylab("") + xlab(" beta coefficients of egfr") + ggstats::geom_stripped_rows()  +  scale_color_paletteer_d("PrettyCols::Bold", name="") + 
  ## alpha: signficant or not
  scale_alpha_discrete(range=c(0.5,1)) 
#ggsave("~/Documents/4CStudy/Plots/forest_betacoefficients_split_diagnose.pdf", width=6,height=12)
```

same plot but separarted for each diagnose, and then but next to each other
```{r}
diagnose_linearmodel_splitted %>%  group_by(analyte)%>% filter(any(FDR < 0.05)) %>% mutate(significant=FDR<0.05) %>% 
ggplot(aes( y=reorder(analyte, -as.numeric(estimate)))) +geom_point(aes(x=estimate,color=diagnose, alpha=significant), shape=16, size=3) +
  geom_linerange(aes(xmin=down, xmax=up,color=diagnose, alpha=significant) ) + theme_classic() + ylab("") + xlab(" beta coefficients of egfr") + ggstats::geom_stripped_rows()  +  scale_color_paletteer_d("PrettyCols::Bold", guide="none") + scale_alpha_discrete(range=c(0.7,1)) + facet_grid(.~ diagnose, scales = "free") + theme(legend.position = "bottom")
#ggsave("~/Documents/4CStudy/Plots/forest_betacoefficients_split_diagnose_nextoteachother.pdf", width=10,height=6)
```

## Contrast

1. Run ANCOVA 3 on simple model: analyte ~ country/diagnose + eGFR , adjusting for eGFR
2. FDR on ANCOVA p-val 
3. on signficiant models (FDR <0.05), post-hoc contrast check  via emmeans

```{r}
library(emmeans)
library(rstatix)
```

### Functions 
ANCOVA: 
function to run ancova for each protein 
data: olink data frame: protein npx values 
predictor: set of clinical variable: egfr and diagnose/country
start,end, positioning of proteins in data
```{r}
run_ancova <- function(data, predictor, group="diagnose2", start,end){
  res <- data.frame(matrix(ncol=2,nrow=0))
  names(res) <- c("analyte", "pval")
  j<-1
  for( i in start:end){
    ## get protein
    y <- data.matrix(data[,i])
    analyte<- names(data)[i]
  
    # Perform ANCOVA 
    formula<- formula(paste("analyte ~ egfr+ ", group))
    anova_result <- predictor %>% mutate(analyte=y) %>% anova_test(formula)
    anova_result <- get_anova_table(anova_result)
   
    # Store the analyte and corresponding p-value
    res[j,] <- c(analyte, anova_result[2, "p"])
    j <- j+1
  
  }
  return(res)}
```


Posthoc Contrast 
function to run the contrast after ancova , with emmeans

data: olink data 
predictors : clinical data (diagnose, country information)
contr: which contrast sets should be investigated : diagnose or country 
maingrp: decided which contrast matrix design is chosen, if NA then one vs all , other wise  maingrp vs minorgrp xx
```{r}
posthoc_contrast_check <- function(data, predictors, contr="diagnose", maingrp=NA){
  ## return dataframe saving name of protein,contrasts, estimates, cis , p-vals,...
  res <- data.frame(matrix(ncol=10,nrow=0))
  names(res) <- c("analyte","contrast", "estimate" ,"SE", "df","lower.CL","upper.CL","t.ratio", "p.value" )
  
  ## for each protein , run the contrasts
  j<-1
  for( i in 1:ncol(data)){
    
    y <- data.matrix(data[,i])
    analyte<- names(data)[i]
    ## linear model: protein ~ group (diagnose or country)
    lm=(lm(scale(analyte)~.,predictors %>% mutate(analyte=y)))
    
    categories_levels <- levels(predictors[,contr])

if (is.na(maingrp)){
      # Create the contrasts automatically for a "one vs all" comparison
      contrast_list <- lapply(categories_levels, function(cat) {
      contrast_vector <- rep(-1 / (length(categories_levels) - 1), length(categories_levels))
      contrast_vector[categories_levels == cat] <- 1
      contrast_vector
      })
      names(contrast_list) <- lapply(categories_levels, function(i) paste(i, "vs all"))
}
else{ ## Turkey/CAKUT vs all other
  contrast_list <- lapply( categories_levels[ categories_levels != maingrp], function(country) {
    contrast_vector <- rep(0, length( categories_levels))  # Start with a vector of 0s
    contrast_vector[ categories_levels == maingrp] <- 1  # Set Turkey/CAKUT to +1
    contrast_vector[ categories_levels == country] <- -1  # Set the other country/diagnose to -1
  
   
    return(contrast_vector)
  })
   # Name the contrast as "Turkey vs Country" or "CAKUT vs Disease"
  names(contrast_list) <- lapply(categories_levels[categories_levels!=maingrp], function(i) paste(maingrp, " vs ", i))
 
}
  # Use emmeans to apply the contrast
  contrast_results <- emmeans::emmeans(lm, specs=contr, contr = contrast_list)
 
  ## save results of contrasts
  ## contrasts: estimate, tukey adjusted pval, CIs,
  ## standardized effect size
  res <- res %>% rbind(data.frame(summary(contrast_results, infer=c(T,T))$contrasts
)  %>% mutate(analyte=analyte,.before=contrast) %>% 
  mutate(effect_size=data.frame(eff_size(contrast_results, sigma=sigma(lm),edf=Inf))$effect.size
)) 
  }
  return(res)

}
```

### Run ANCOVA 
run ancova protein ~ egfr + diagnose/country 
and save the proteins with signficant association with diagnose/country (FDR <0.05)
```{r}
##### diagnose
signifi_analytes_diangose_ancova<- run_ancova(data=olink_data %>% filter(patid != pat_withoutegfr) %>% arrange(patid), predictor = data.frame(select(BL_data_bpoffice,egfr,diagnose2)), group="diagnose2", 82,158)
## FDR filter
signifi_analytes_diangose_ancova<-  signifi_analytes_diangose_ancova %>% filter(p.adjust(pval,"fdr") < 0.05) %>% pull(analyte)
```

```{r}
#### country
signifi_analytes_country_ancova<- run_ancova(data=olink_data %>% filter(patid != pat_withoutegfr) %>% arrange(patid), predictor = data.frame(select(BL_data_bpoffice,egfr,country)), group = "country", 82,158)
## fdr filter
signifi_analytes_country_ancova<-signifi_analytes_country_ancova%>% filter(p.adjust(pval,"fdr") < 0.05) %>% pull(analyte)
```
### Run Contrast
run contrasts for ancova signifcant associations
contrast design: CAKUT and Turkey as main groups
```{r}
posthoc_contras_result <- posthoc_contrast_check(olink_data %>% filter(patid != pat_withoutegfr) %>% arrange(patid) %>% select(signifi_analytes_diangose_ancova), data.frame(BL_data_bpoffice  %>% arrange(patid) %>% select(diagnose2,egfr,age,sex,bmisds) %>% mutate(diagnose2=factor(diagnose2))), contr="diagnose2", maingrp="CAKUT")

country_posthoc_contras_result <- posthoc_contrast_check(olink_data %>% filter(patid != pat_withoutegfr) %>% arrange(patid) %>% select( signifi_analytes_country_ancova ), data.frame(BL_data_bpoffice %>%  arrange(patid) %>% select(country,egfr,age,sex,bmisds) %>% mutate(country=factor(country))), contr = "country", maingrp="Turkey")
```

### Visualize 
as heatmap , color is the standardized effect size, and star annotated tukey adj. signficant pvalues
```{r}
library(ComplexHeatmap)
visualize_contrast_res <- function(posthoc_contras_result){

# plotting only proteins with at least one signifcant contrast
    
## effect size coloring
heatmap_data <- posthoc_contras_result %>% group_by(analyte) %>% filter(any(p.value < 0.05) ) %>% ungroup()%>%mutate( analyte=str_replace(analyte,"^rawscale_","")) %>%
  select(analyte, contrast, effect_size) %>%
  pivot_wider(names_from = contrast, values_from = effect_size, values_fill = 0) %>%
  column_to_rownames("analyte") %>%
  as.matrix()

## star annotations
signif_matrix <- posthoc_contras_result %>% group_by(analyte) %>% filter(any(p.value < 0.05) ) %>% ungroup() %>%
 mutate( analyte=str_replace(analyte,"^rawscale_","")) %>%
  select(analyte, contrast, p.value) %>%
  pivot_wider(names_from = contrast, values_from = p.value, values_fill = 0) %>%
  column_to_rownames("analyte") %>%
  as.matrix()
 

### labels 
labels <- read_excel("~/Projects/4cstudy/ClinicalData/analyte_abbrev_HB.xlsx") 
labels <- labels %>% filter(analyte %in% rownames(heatmap_data)) %>% arrange(factor(analyte,levels= rownames(heatmap_data))) %>% pull(name)

rownames(heatmap_data) <- labels
rownames(signif_matrix) <- labels
# Generate heatmap with annotations
heatmap <- Heatmap(
  heatmap_data,
  name = "Effect Size",
  col = circlize::colorRamp2(c(min(heatmap_data), 0, max(heatmap_data)), rev(c("#075AFF", "#FFFFCC", "#FF0000"))),
  cluster_rows = TRUE,
  cluster_columns = F,
  show_row_dend = F,
  show_column_dend = F,
  cell_fun = function(j, i, x, y, width, height, fill) {
   if(signif_matrix[i, j] < 0.001) {
        grid.text("***", x, y)
    } else if(signif_matrix[i, j] < 0.01) {
        grid.text("**", x, y)
    }
    else if(signif_matrix[i, j] < 0.05) {
        grid.text("*", x, y)
    }
    else {
        grid.text("", x, y)
    }
  }
) #

return(heatmap)

}
```

```{r}
## diagnose heatmap
visualize_contrast_res(posthoc_contras_result %>% mutate(contrast=if_else(endsWith(contrast,"loss"), "CAKUT vs postAKI/HUS",contrast))) 

## country heatmap
visualize_contrast_res(country_posthoc_contras_result)

## save contrast results
posthoc_contras_result %>% mutate(contrast=if_else(endsWith(contrast,"loss"), "CAKUT vs postAKI/HUS",contrast), 
                                  analyte=str_remove(analyte,"rawscale_")) %>% write.table(file="~/Projects/4cstudy/Tables/diagnose_contrast_results.csv", sep = ",",row.names = F)

country_posthoc_contras_result %>% mutate(analyte=str_remove(analyte,"rawscale_")) %>% write.table(file="~/Projects/4cstudy/Tables/country_contrast_results.csv", sep = ",",row.names = F)

```