---
title: "olink preprocessing, Qualtiy Control and PCA"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(tibble)
library(ggpubr)
library(readxl)
library(paletteer)
```

## 0. Read files 
```{r message=FALSE, warning=FALSE}
setwd("~/Projects/4cstudy")
olink_file <- read.csv2("~/Projects/4cstudy/ClinicalData/20241112_cohort_4C.csv", sep=",", dec=".")
olink_metafile <- read_xlsx("~/Projects/4cstudy/ClinicalData/20240917_Olink_layout_AG_Wilck.xlsx") %>% filter(cohort=="4C")
LOD_file <- read.csv2("~/Projects/4cstudy/ClinicalData/analytes_lod.csv", sep=",", dec=".")


olink_file %>% filter(repeated=="yes") %>% distinct(pat_id)
```

## 1. Data Cleaning 

### Repeated Measurements
```{r}
length(unique(olink_file$pat_id))
pat_with_olink <- unique(olink_file$pat_id)

## adding all with no repeated measuremtns 
olink_data<-olink_file %>% group_by(pat_id) %>% filter(all(repeated=="no")) 

## 10 patients with  repeated measurements, with changed sample nr , taken the repeated measurement as original consists solely on 0 for all analytes
filteringout_samples <- olink_file %>% group_by(pat_id) %>% filter(any(repeated=="yes"))  %>% filter(!startsWith(unchanged_sample_nr,"W")) %>% distinct(unchanged_sample_nr) %>% pull(unchanged_sample_nr)

olink_data <- rbind( olink_data , olink_file %>% group_by(pat_id) %>% filter(any(repeated=="yes"))  %>% filter(startsWith(unchanged_sample_nr,"W")))

# change format of olink_data
olink_data <- olink_data %>% select(unchanged_sample_nr, sample_nr ,pat_id, analytes, value, qc_by_them)  %>% pivot_wider(names_from=analytes, values_from = value)
```

### Warning Flag

38 patients with warning 
```{r}
nrow( olink_data %>% filter(qc_by_them=="WARNING") %>% group_by(pat_id) %>% distinct(pat_id))
nrow( olink_data %>% filter(qc_by_them=="WARNING") %>% group_by(pat_id) %>% distinct(pat_id))

```
### Adding Plate 

```{r}
olink_data<- olink_metafile %>% filter(cohort=="4C") %>% select(sample_nr=sampleNr, plate) %>% full_join(olink_data %>% mutate(sample_nr=as.character(sample_nr)),by="sample_nr")


olink_file <- olink_metafile %>% filter(cohort=="4C") %>% select(sample_nr=sampleNr, plate) %>% full_join(olink_file %>% mutate(sample_nr=as.character(sample_nr)),by="sample_nr")

plate_sizes <- olink_file %>%group_by(plate) %>% summarise(n_samples=n_distinct(pat_id))

plate_sizes
```


## 2 LOD 

### sample based
all patients have some analytes under LOD 
seemed to occur on every plate
```{r}
#LOD_file

olink_file %>% left_join(LOD_file %>% rename(analytes=anlaytes), by= "analytes")  %>% filter(!unchanged_sample_nr %in% filteringout_samples) %>% mutate(underLOD= value < MaxLOD) %>% filter(underLOD) %>% group_by(pat_id,plate) %>% summarise(n_underlod=n_distinct(analytes)) %>% ggplot(aes(y=n_underlod,x=as.character(plate))) + geom_boxplot()
```


max 22% missing analytes, so under the 0.7 treshold 
```{r}
olink_file %>% left_join(LOD_file %>% rename(analytes=anlaytes), by= "analytes")  %>% filter(!unchanged_sample_nr %in% filteringout_samples) %>% 
mutate(underLOD= value < MaxLOD) %>% filter(underLOD) %>% group_by(unchanged_sample_nr) %>% summarise(n_underlod=n_distinct(analytes), perc=n_underlod/92,qc_by_them) %>% arrange(desc(n_underlod)) %>% distinct()

## samples with enough analytes above LOD 
olink_file %>% left_join(LOD_file %>% rename(analytes=anlaytes), by= "analytes")  %>% filter(!unchanged_sample_nr %in% filteringout_samples) %>% 
mutate(underLOD= value < MaxLOD) %>% filter(underLOD) %>% group_by(unchanged_sample_nr) %>% summarise(n_underlod=n_distinct(analytes), perc=n_underlod/92,qc_by_them) %>% arrange(desc(n_underlod)) %>% pull(perc) %>% summary(.)
```

### analytes based
```{r}
analytes_occurrence <- olink_file %>% left_join(LOD_file %>% rename(analytes=anlaytes), by= "analytes")  %>% filter(!unchanged_sample_nr %in% filteringout_samples) %>% mutate(underLOD= value < MaxLOD) %>% filter(underLOD) %>% group_by(analytes) %>% count() %>% mutate(perc=n/684) %>% arrange(desc(n)) 

analytes_occurrence

filterout_analytes <- analytes_occurrence %>% filter(perc > 0.7) %>% pull(analytes)


## whether this analytes are missing by chance or systematically (ckd status?), see under 5)
analytes_tobetested <- analytes_occurrence %>% filter(perc > 0.7 & perc <0.9) %>% pull(analytes)

critical_analytes_absent_df <- olink_file%>% left_join(LOD_file %>% rename(analytes=anlaytes), by= "analytes")  %>% filter(!unchanged_sample_nr %in% filteringout_samples) %>% mutate(underLOD= value < MaxLOD) %>% filter(analytes %in% analytes_tobetested) %>% select(pat_id,analytes,underLOD) %>% pivot_wider(names_from=analytes,values_from=underLOD)
```

how many analytes are not measured on the plates 
seem to be also similar on the plates

percentage of missing for each analyte on each plate

```{r}
olink_file %>% left_join(LOD_file %>% rename(analytes=anlaytes), by= "analytes")  %>% filter(!unchanged_sample_nr %in% filteringout_samples) %>% mutate(underLOD= value < MaxLOD) %>% filter(underLOD) %>% group_by(analytes,plate) %>% count(,Missing.Data.freq.) %>% arrange(analytes) %>% left_join(plate_sizes, by="plate") %>% mutate(perc=n/n_samples)
```

total number of missing analytes per plate 
```{r}
olink_file %>% left_join(LOD_file %>% rename(analytes=anlaytes), by= "analytes")  %>% filter(!unchanged_sample_nr %in% filteringout_samples) %>% mutate(underLOD= value < MaxLOD) %>% filter(underLOD) %>% group_by(analytes,plate) %>% count(,Missing.Data.freq.) %>% arrange(analytes)  %>% #filter(!analytes %in% filterout_analytes) %>% 
ungroup()%>% count(plate)

```
### Warning flag

warning not based on under LOD, as only 43 of them are from samples flagged with warning 
```{r}
nrow(olink_file %>% left_join(LOD_file %>% rename(analytes=anlaytes), by= "analytes")  %>% filter(!unchanged_sample_nr %in% filteringout_samples) %>% mutate(underLOD= value < MaxLOD) %>% filter(underLOD) %>% filter(!analytes %in% filterout_analytes) %>% filter(qc_by_them =="WARNING"))

## warning not based on lod 
```

```{r}
ggplot(olink_file %>% filter(qc_by_them!="PASS" & !unchanged_sample_nr %in% filteringout_samples) , aes(x=reorder(pat_id,value,mean), y=value)) + geom_boxplot()

## enough values above LOD
olink_file%>% left_join(LOD_file %>% rename(analytes=anlaytes), by= "analytes")  %>% mutate(underLOD= value < MaxLOD) %>% filter(qc_by_them!="PASS" & !unchanged_sample_nr %in% filteringout_samples) %>% group_by(pat_id) %>%  filter(underLOD) %>% count(underLOD)
```


```{r}
ggplot(rbind(olink_file %>% filter(qc_by_them=="PASS") %>% group_by(unchanged_sample_nr) %>% distinct(unchanged_sample_nr)  %>% pull(unchanged_sample_nr) %>% sample(.,10)  %>% { filter(olink_file, unchanged_sample_nr %in% .) }   ,
           olink_file %>% filter(qc_by_them!="PASS") %>% filter(!unchanged_sample_nr %in% filteringout_samples)%>% group_by(unchanged_sample_nr) %>% distinct(unchanged_sample_nr)  %>% pull(unchanged_sample_nr) %>% sample(.,10)  %>% { filter(olink_file, unchanged_sample_nr %in% .) } )  , 
       aes(x=interaction(unchanged_sample_nr,qc_by_them), y=value, color=qc_by_them)) + geom_boxplot() + theme_classic()

olink_file %>% filter(!unchanged_sample_nr %in% filteringout_samples)%>% ggplot(aes(x=qc_by_them, y=value)) + geom_boxplot() + stat_compare_means()

```


## 3 Imputation

taken the square root of max LOD
issue: since value and LOD are log transformed, can not take square root for negative LODs
therefore reverse log transformation and taking the square root and log transform then again 

rawscale_value are those imputated values 
logscale_value taken square root of log maxLOD, hence some with missing value 
```{r}
## add LOD values for each analyte, create imputation on raw and on log value of LOD
olink_file <- olink_file %>% left_join(LOD_file %>% rename(analytes=anlaytes), by= "analytes")  %>% filter(!unchanged_sample_nr %in% filteringout_samples & !analytes %in% filterout_analytes)  %>% mutate(underLOD=value < MaxLOD) %>%
mutate( expLOD=exp(MaxLOD)) %>% 
mutate(imput_raw_value=case_when(underLOD ~ log(sqrt(expLOD)), T ~ value))  %>%
mutate(imput_log_value = case_when(underLOD ~ sqrt(MaxLOD), T ~ value))

## change format of olink file 
olink_data <- olink_file %>% select(unchanged_sample_nr,plate,qc_by_them, pat_id, analytes, unchanged=value, rawscale=imput_raw_value, logscale=imput_log_value) %>% pivot_wider(
    names_from = analytes,
    values_from = c(unchanged, rawscale, logscale),
    names_glue = "{.value}_{analytes}"
  )

## input for all other analysis
#write.table(olink_data, "~/Documents/4CStudy/Tables/OLINK_data.tsv", row.names=F, quote=F,na="NA", sep="\t")
```


## 4 PCA
### functions
```{r}
## calvcualtes the prinicpial components of the given data
calculate_PCA <- function(data){
pca_result <- prcomp(data %>% dplyr::select(-pat_id), scale. = F )  
# Step 2: Prepare PCA results for plotting
# Get the principal components (PCs) and bind with patient IDs
pca_df <- as.data.frame(pca_result$x)
pca_df$pat_id <-data   %>% pull(pat_id)

return(pca_df)
}

# get the variance covered by the first and second principal component, for axis description
calculate_PCs <-function(data){
pca_result <- prcomp(data %>% select(-pat_id), scale. = F )  

variance_explained <- summary(pca_result)$importance[2, 1:2] * 100  # Convert to percentage
pc1_var <- round(variance_explained[1], 2)
pc2_var <- round(variance_explained[2], 2)

return(c(pc1_var,pc2_var))
}
```

### General Checks
any clustering based on plate and quality control flags ? 

```{r}
# calculate PCs with the "raw" imputated npx values, and then join with plate and qc info data for coloring
raw_pca_df <- calculate_PCA(olink_data %>% select(pat_id, starts_with("raw"))) %>% 
  right_join(olink_data %>% select(pat_id,plate,qc_by_them),., by="pat_id")

# get the variance of first,second PCs
raw_pcs <- calculate_PCs(olink_data %>% select(pat_id, starts_with("raw")))
```


```{r}
ggplot(raw_pca_df, aes(x = PC1, y = PC2, color = qc_by_them)) +
  geom_point(size = 3) +
  theme_minimal() +  scale_color_viridis_d()  +
  labs(
    x = paste0("PC 1 (", raw_pcs[1], "% variance)"),
    y = paste0("PC 2 (", raw_pcs[2], "% variance)")
  )

ggplot(raw_pca_df, aes(x = PC1, y = PC2, color = as.character(plate))) +
  geom_point(size = 3) +
  theme_minimal() +  scale_color_viridis_d()  +
  labs(
    x = paste0("PC 1 (", raw_pcs[1], "% variance)"),
    y = paste0("PC 2 (", raw_pcs[2], "% variance)")
  )
```


### colored by CKD

```{r include=T}
## clinical data at study entrance (BL = baseline), adding information within ggplot
BL_data_bpoffice <-read.csv2("~/Projects/4cstudy/Tables/BL_data_bpoffice.csv",  sep = ",",header=T)
```

for better visualization,excluding 8 patients of ckd 2 stage (> 60 eGFR)
```{r}
ggplot(
  raw_pca_df %>% 
    rename(patid = pat_id) %>% 
    right_join(BL_data_bpoffice, ., by = "patid") %>% 
    mutate(egfr = as.numeric(egfr)) %>%  filter(egfr < 60), 
  
) + # Scatterplot points
 geom_point(aes(x = PC1, y = PC2, color = egfr), shape=16, size = 2.5) +
  # Add median points for each group
  geom_point(
    data = raw_pca_df %>% rename(patid = pat_id) %>% right_join(BL_data_bpoffice, ., by = "patid") %>%
      mutate(egfr = as.numeric(egfr)) %>%  filter(egfr < 60) %>%  group_by(ckd1) %>% 
      summarise(  median_PC1 = median(PC1, na.rm = TRUE), median_PC2 = median(PC2, na.rm = TRUE)),
    aes(x = median_PC1, y = median_PC2, fill=ckd1),  shape=24 ,size=5,color="white"
  ) +
  # Classic theme and color scale
  theme_classic() +  
  scale_color_viridis_c() +
  scale_fill_viridis_d(direction=-1)+
  # Axis labels
  labs(
    x = paste0("PC 1 (", raw_pcs[1], "% variance)"),
    y = paste0("PC 2 (", raw_pcs[2], "% variance)")
  )

#ggsave("~/Documents/4CStudy/Plots/PCA_egfr_CKDmedian.pdf", width=7,height=5)
```

all patients, from ckd stage 2 to 5
```{r}
ggplot(raw_pca_df %>% rename(patid=pat_id)%>% left_join(BL_data_bpoffice, by="patid") %>% filter(!is.na(ckd1)), aes(x = PC1, y = PC2, color = ckd1)) +
  geom_point(size = 3) +
  theme_minimal() +  scale_color_viridis_d()  +
  labs(
    x = paste0("PC 1 (", raw_pcs[1], "% variance)"),
    y = paste0("PC 2 (", raw_pcs[2], "% variance)")
  )
```

### colored by Diagnose
adding info from BL table and changing names of diagnose for "Nephloss"
```{r}
ggplot(raw_pca_df %>% rename(patid=pat_id)%>% left_join(BL_data_bpoffice, by="patid") %>% mutate(diagnose2=case_when(diagnose2=="Nephloss"~"postAKI/HUS",T~diagnose2)) %>%
  mutate(diagnose2=factor(diagnose2, levels=rev(c("Other", "postAKI/HUS", "Glomerulopathies","Tubint", "CAKUT")))), aes(x = PC1, y = PC2, color = diagnose2)) +
  geom_point(size = 3) +
  theme_classic() +  scale_color_paletteer_d("PrettyCols::Bold", name="")  +
  labs(
    x = paste0("PC 1 (", raw_pcs[1], "% variance)"),
    y = paste0("PC 2 (", raw_pcs[2], "% variance)")
  ) + geom_point(
    data = raw_pca_df %>% rename(patid = pat_id) %>% right_join(BL_data_bpoffice, ., by = "patid") %>%
      mutate(diagnose2=case_when(diagnose2=="Nephloss"~"postAKI/HUS",T~diagnose2)) %>%
  mutate(diagnose2=factor(diagnose2, levels=rev(c("Other", "postAKI/HUS", "Glomerulopathies","Tubint", "CAKUT")))) %>%
       group_by(diagnose2) %>% 
      summarise(  median_PC1 = median(PC1, na.rm = TRUE), median_PC2 = median(PC2, na.rm = TRUE)),
    aes(x = median_PC1, y = median_PC2, fill=diagnose2),  shape=24 ,size=5, color="white"
  ) +
  # Classic theme and color scale
  theme_classic() +  
  paletteer::scale_fill_paletteer_d("PrettyCols::Bold", name="")+
  # Axis labels
  labs(
    x = paste0("PC 1 (", raw_pcs[1], "% variance)"),
    y = paste0("PC 2 (", raw_pcs[2], "% variance)")
  )

#ggsave("~/Documents/4CStudy/Plots/PCA_disease_withmedian.pdf", width=7,height=5)
```

### colored by site
colored by country of site (turkey,france,germany,...)
```{r}
ggplot(raw_pca_df %>% rename(patid=pat_id)%>% left_join(BL_data_bpoffice, by="patid") , aes(x = PC1, y = -PC2, color = as.character(country))) +
  geom_point(size = 3) +
  theme_classic() +  paletteer::scale_color_paletteer_d("MoMAColors::Klein", name="") +
  labs(
    x = paste0("PC 1 (", raw_pcs[1], "% variance)"),
    y = paste0("PC 2 (", raw_pcs[2], "% variance)")
  ) +
  geom_point(
    data = raw_pca_df %>% rename(patid = pat_id) %>% right_join(BL_data_bpoffice, ., by = "patid") %>%
       group_by(country) %>% 
      summarise(  median_PC1 = median(PC1, na.rm = TRUE), median_PC2 = median(PC2, na.rm = TRUE)),
    aes(x = median_PC1, y = median_PC2, fill=country),  shape=24 ,size=5, color="white"
  ) +
  # Classic theme and color scale
  theme_classic() +  
  paletteer::scale_fill_paletteer_d("MoMAColors::Klein", name="")+
  # Axis labels
  labs(
    x = paste0("PC 1 (", raw_pcs[1], "% variance)"),
    y = paste0("PC 2 (", raw_pcs[2], "% variance)")
  )
#ggsave("~/Documents/4CStudy/Plots/PCA_country_withmedian.pdf", width=7,height=5)
```

### colored by proteinuria
```{r}
raw_pca_df %>% rename(patid=pat_id)%>% left_join(BL_data_bpoffice, by="patid") %>% filter(!is.na(alburie))%>%
  mutate(alburie=as.numeric(alburie)) %>%mutate(acr_grp= case_when(alburie< 30 ~"A1", alburie>300~"A3",T~"A2") ) %>%
ggplot( ., aes(x = PC1, y = -PC2, color = log(alburie+0.1))) +
  geom_point(size = 3) +
  theme_classic() +
  geom_point(
    data = raw_pca_df %>% rename(patid = pat_id) %>% right_join(BL_data_bpoffice, ., by = "patid") %>% filter(!is.na(alburie))%>%
      mutate(acr_grp= case_when(as.numeric(alburie)< 30 ~"A1",as.numeric(alburie)>300~"A3",T~"A2") ) %>%
    # mutate(acr_grp=as.character(ntile(alburie,6))) %>%
       group_by(acr_grp) %>% 
      summarise(  median_PC1 = median(PC1, na.rm = TRUE), median_PC2 = -median(PC2, na.rm = TRUE)),
    aes(x = median_PC1, y = median_PC2, fill=acr_grp),  shape=24 ,size=5, color="white") +
  # Classic theme and color scale
  theme_classic() +  
  scale_color_viridis_c() + scale_fill_viridis_d()+
  # Axis labels
  labs(
    x = paste0("PC 1 (", raw_pcs[1], "% variance)"),
    y = paste0("PC 2 (", raw_pcs[2], "% variance)")
  )


```
### colored by sex 
```{r}
raw_pca_df %>% rename(patid=pat_id)%>% left_join(BL_data_bpoffice, by="patid") %>% filter(!is.na(sex)) %>%
ggplot( ., aes(x = PC1, y = -PC2, color = sex)) +
  geom_point(size = 3) +
  theme_classic() +
  geom_point(
    data = raw_pca_df %>% rename(patid = pat_id) %>% right_join(BL_data_bpoffice, ., by = "patid")   %>% filter(!is.na(sex)) %>%
       group_by(sex) %>% 
      summarise(  median_PC1 = median(PC1, na.rm = TRUE), median_PC2 = -median(PC2, na.rm = TRUE)),
    aes(x = median_PC1, y = median_PC2, fill=sex),  shape=24 ,size=5, color="white") +
  # Classic theme and color scale
  theme_classic() +  
  # Axis labels
  labs(
    x = paste0("PC 1 (", raw_pcs[1], "% variance)"),
    y = paste0("PC 2 (", raw_pcs[2], "% variance)")
  ) +
 paletteer::scale_fill_paletteer_d("PrettyCols::Bold", name="") + paletteer::scale_color_paletteer_d("PrettyCols::Bold", name="")


```

### colored by age
```{r}
median<- median(BL_data_bpoffice %>% filter(!is.na(age_in_years_at_vdate)) %>% 
         mutate(age_in_years_at_vdate=as.numeric(age_in_years_at_vdate)) %>%
         pull(age_in_years_at_vdate))


raw_pca_df %>% rename(patid=pat_id)%>% left_join(BL_data_bpoffice, by="patid") %>% filter(!is.na(age_in_years_at_vdate)) %>%  mutate(age_in_years_at_vdate=as.numeric(age_in_years_at_vdate)) %>%
ggplot( ., aes(x = PC1, y = -PC2, color = age_in_years_at_vdate)) +
  geom_point(size = 3) +
  theme_classic() +
  geom_point(
    data = raw_pca_df %>% rename(patid = pat_id) %>% right_join(BL_data_bpoffice, ., by = "patid")   %>% filter(!is.na(age_in_years_at_vdate)) %>%
       mutate( age_above= if_else(median(age_in_years_at_vdate) <age_in_years_at_vdate, "yes", "no")) %>% group_by(age_above)%>%
      summarise(  median_PC1 = median(PC1, na.rm = TRUE), median_PC2 = -median(PC2, na.rm = TRUE)),
    aes(x = median_PC1, y = median_PC2, fill=age_above),  shape=24 ,size=5, color="white") +
  # Classic theme and color scale
  theme_classic() +  
  # Axis labels
  labs(
    x = paste0("PC 1 (", raw_pcs[1], "% variance)"),
    y = paste0("PC 2 (", raw_pcs[2], "% variance)"),
    color="age in years",
    fill=paste0("age < median age of ", round(median,2))
  ) +
 scale_colour_viridis_c() + scale_fill_viridis_d()



```

## 5 Missing Proteins
7 proteins with a missing rate of 70% up to 90% 
by chance or systemtically (kidney damage?)
defined under 2) analytes based
have a look on whether those proteins are missing more or less in certain ckd stages
```{r}
critical_analytes_absent_df %>% rename(patid=pat_id) %>% left_join(BL_data_bpoffice,.,by="patid") %>% group_by(ckd1) %>%   summarise(across(analytes_tobetested, ~sum(.x, na.rm = TRUE)/n())) %>% ungroup() %>% gather(-ckd1,key="protein", value="value" ) %>% 
  ggplot(aes(x=protein,fill=ckd1, y=value)) + geom_bar(stat="identity", position="dodge") + theme_classic() +   scale_fill_paletteer_d("PrettyCols::Bold", name="CKD stage") + ylab("Percentage of missing meausrements") + xlab("")
```