---
title: "Protein Network and I-score Subcohorts"
output:
  html_document:
    df_print: paged
---

more detailed description of the subcohorts (inflammation degree)
```{r include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(readxl)
library(paletteer)
library(metadeconfoundR)
library(fastDummies)
library(stringr)
library(tibble)
library(igraph)
library(ggraph)
theme_classic_modified <- function(){
  theme_classic() %+replace%
    theme(strip.placement="outside", 
                           strip.text.y.left = element_text(angle = 0,size=11),
                           strip.background.y=element_rect(color = NA), 
                           strip.background.x=element_rect(color = NA),
                           legend.position = "none") 
}
```

functions
```{r}
plot_distribution <- function(data,x="i_score_80_cat",fill,savepath=NULL){
## chi sqaure test
 contingency_table <- table(data %>% pull(!!sym(x)),  data %>% pull(!!sym(fill)))
 print(contingency_table)
chisqrtest <- chisq.test(contingency_table)

plot <- data%>% dplyr::count(!!sym(x),!!sym(fill)) %>% group_by(!!sym(x)) %>% mutate(perc=n/sum(n)) %>% 
  ggplot(aes(x=!!sym(x),y=perc,fill=!!sym(fill))) + 
  geom_col(position=position_stack()) + 
  geom_text(aes(label=n), position = position_stack(vjust =0.5), size =3.5, color="white") + 
  theme_classic() + scale_fill_viridis_d() + 
  labs(x="", y="", subtitle=paste0("chisquare test: X: ",round(chisqrtest$statistic,2),", p-val: ", round( chisqrtest$p.value,2)))

if(!is.null(savepath)){
  ggsave(paste0(savepath,"bar_distrib_",x,"_",fill,".pdf"),plot, width = 5.5,height=4.6)
}
return(plot)
}
```

## 0 Data
```{r}
### defined inflammation score with a cutoff of 80th, based on egfr and proteinuria -> 660 patients 
coxdata <-  read.csv("~/Projects/4cstudy/Tables/survival_data.csv",  sep = ",",header=T)
coxdata <- coxdata %>% mutate(i_score_80_cat= case_when(i_score==0~"0",i_score==1 | i_score==2~"1-2",T~"3-4"))

olink_data <- read.csv2("~/Projects/4cstudy//Tables/OLINK_data.csv",  sep = ",",header=T) %>% arrange(pat_id) %>% rename(patid=pat_id) %>% mutate(across(starts_with("rawscale"), as.numeric))

### BL_data_bpoffice -> baseline table based on investigation file , more information than investigation_file, end product of the Clinical Data Preparation script, including medication, hypertension,...
BL_data_bpoffice <-read.csv2("~/Projects/4cstudy/Tables/BL_data_bpoffice.csv",  sep = ",",header=T) %>% filter(patid != 737) %>% mutate(across(everything(), ~ type.convert(.x, as.is = TRUE)))

labels <- read_excel("~/Projects/4cstudy/ClinicalData/analyte_abbrev_HB.xlsx") 
```

```{r}
coxdata <- coxdata %>% mutate(i_score_80_cat =case_when(i_score_80==0~"0",i_score_80>=1 & i_score_80<3~ "1-2",T~"3-4"))
```

## 1  Continuous Variables

table with mean, sd, and number of missing values
```{r} 
coxdata%>% group_by(i_score_80_cat) %>% 
summarise(across(c(BLage,egfr,alburie,salb,hb,bicar,phos,syssds,diasds,bmisds), list(mean = mean, sd = sd),na.rm=T)) %>% left_join(coxdata%>% group_by(i_score_80_cat) %>% 
summarise(across(c(BLage,egfr,alburie,salb,hb,bicar,phos,syssds,diasds,bmisds), ~sum(is.na(.)),.names = "{.col}_NAs"))   )   
```

### metadeconfoundR
running metadeconfoundR ( confounding not relevant here, but is way to iterate over the naive association and plotting heatmap )
```{r}
featuredata <- coxdata %>%left_join(BL_data_bpoffice%>% dplyr::select(patid,lvmi,imt_sds_a,pwv_sds_a,sum_AHT,sum_medication))%>%
  dplyr::select(patid,survival_time, BLage,bmisds,alburie,salb,hb,bicar,phos,syssds,diasds,sum_AHT,sum_medication,starts_with("residual"))   %>% dplyr::select(-patid)

labels_heatmap <- data.frame(machine= names(featuredata), human= c(names(featuredata)[1:12],paste0("res. ", labels$name )))


results<- MetaDeconfound(
  featureMat = featuredata, 
  metaMat = coxdata%>% dummy_cols("i_score_80_cat",remove_first_dummy = T,remove_selected_columns = T) %>% dplyr::select(starts_with("i_score_80"),high_80_residual_tnfrsf9,high_80_residual_cd40,high_80_residual_cx3cl1,high_80_residual_pd_l1,egfr) , returnLong = T, logLevel = "ERROR"
)
```




```{r}
BuildHeatmap(results,reOrder = "feat",featureNames=labels_heatmap) + labs(x="",y="",title="")
```
### boxplots
```{r}
coxdata %>% pivot_longer(c(bmisds,salb,phos,hb), names_to="key",values_to="val") %>% ggplot(aes(x=i_score_80_cat,y=val,fill=i_score_80_cat)) + 
  geom_boxplot()+ facet_wrap(~key,scales="free_y") +  scale_fill_manual(values=c( "#FFD353FF","#FFB242FF","#BB292CFF"), name="i-score 80") + theme_classic_modified() + ggpubr::stat_compare_means(size=3) + ylab("") + xlab("i-score")
```





## 2 Categorical Variables
```{r}
path <- "~/Projects/4cstudy/Plots/Figure5/"
```

sex
```{r}
BL_data_bpoffice %>% right_join(coxdata%>%dplyr::select(patid,i_score_80_cat)) %>% 
dplyr::count(i_score_80_cat,sex) %>% group_by(i_score_80_cat) %>% mutate(perc=n/sum(n))
plot_distribution(data=BL_data_bpoffice %>% right_join(coxdata %>% mutate(i_score_80=as.character(i_score_80_cat))%>%dplyr::select(patid,i_score_80)) , fill="sex", x="i_score_80", savepath = path)
```


### Hypertension


```{r}
#on AHT or not
BL_data_bpoffice %>% right_join(coxdata%>%dplyr::select(patid,i_score_80_cat,i_score_80)) %>% 
dplyr::count(i_score_80_cat,AHT) %>% group_by(i_score_80_cat) %>% mutate(perc=n/sum(n))
# hypertensive
BL_data_bpoffice %>% right_join(coxdata%>%dplyr::select(patid,i_score_80_cat,i_score_80)) %>% 
dplyr::count(i_score_80_cat,hyper_class) %>% group_by(i_score_80_cat) %>% mutate(perc=n/sum(n))
```

```{r}
## hypertension group: hypertension (yes, no), treated (yes,no)
plot_distribution(data=BL_data_bpoffice %>% right_join(coxdata%>%dplyr::select(patid,i_score_80_cat)) , fill="hyper_treat_group",savepath = path) 

## number of anti hypertensive drugs
plot_distribution(data=BL_data_bpoffice %>% mutate(sum_AHT=as.character(sum_AHT)) %>% right_join(coxdata%>%dplyr::select(patid,i_score_80_cat)) , fill="sum_AHT",savepath = path)
```


```{r}
## cvd manifestations
plot_distribution(data=BL_data_bpoffice %>% group_by(patid) %>% mutate(sum_cvd= as.character(sum(lvmi_critical+imt_sds_a_above2+pwv_sds_a_above2,na.rm=T))) %>% ungroup() %>% right_join(coxdata%>%dplyr::select(patid,i_score_80_cat)) , fill="sum_cvd",savepath = path) 
```
```{r}
## all type of medication
plot_distribution(data=BL_data_bpoffice %>% mutate(sum_medication=if_else(sum_medication<=2,as.character(sum_medication),"3 or more")) %>% right_join(coxdata%>%dplyr::select(patid,i_score_80_cat)) , fill="sum_medication",savepath = path) 
```
```{r}
## RAS inhibitor
plot_distribution(data=BL_data_bpoffice %>% filter(!is.na(medigr_RAS.inhibitors)) %>% mutate(RAS=as.character(medigr_RAS.inhibitors)) %>% inner_join(coxdata%>%dplyr::select(patid,i_score_80_cat)) , fill="RAS",savepath = path) 
```
```{r}
## immunsuppressive
plot_distribution(data=BL_data_bpoffice %>% filter(!is.na(medigr_Immunsuppressiva)) %>% mutate(immunsuppressiva=as.character(medigr_Immunsuppressiva)) %>% inner_join(coxdata%>%dplyr::select(patid,i_score_80_cat)) , fill="immunsuppressiva",savepath = path) 
```


### underlying disease

```{r message=FALSE, warning=FALSE}
plot_distribution(data=BL_data_bpoffice %>% right_join(coxdata%>%dplyr::select(patid,i_score_80_cat)) , fill="diagnose2") + scale_fill_paletteer_d("PrettyCols::Bold", name="")

plot_distribution(data=BL_data_bpoffice %>% right_join(coxdata%>%dplyr::select(patid,i_score_80) %>%mutate(i_score_80=as.character(i_score_80))) , x="diagnose2",fill="i_score_80", savepath = path) 

```

### country 

```{r message=FALSE, warning=FALSE}
plot_distribution(data=BL_data_bpoffice %>% right_join(coxdata%>%dplyr::select(patid,i_score_80_cat)) , fill="country",savepath = path) 
```


##  3 Protein Profile
### PCA 
```{r}
x<- olink_data%>% filter(patid %in% coxdata$patid) %>% dplyr::select(patid,starts_with("raw")) %>% arrange(patid) %>% column_to_rownames("patid")

pca_df <- prcomp(x , scale. = F ) 
variance_explained <- summary(pca_df)$importance[2, 1:2] * 100  # Convert to percentage
pc1_var <- round(variance_explained[1], 2)
pc2_var <- round(variance_explained[2], 2)

pca_df <- as.data.frame(pca_df$x)
pca_df$patid <-as.numeric(rownames(x))


pca_df %>%  left_join(coxdata) %>% ggplot(aes(x=PC1,y=PC2)) + geom_point(aes(color=i_score_80_cat)) + theme_classic() + geom_point(
    data = pca_df  %>% right_join(coxdata, ., by = "patid") %>%   
      group_by(i_score_80_cat) %>%  
      summarise(  median_PC1 = median(PC1, na.rm = TRUE), median_PC2 = median(PC2, na.rm = TRUE)),  
    aes(x = median_PC1, y = median_PC2, fill=i_score_80_cat),  shape=24 ,size=5,color="white" )+ ggtitle("PCA npx protein values") + 
  labs(
    x = paste0("PC 1 (", pc1_var, "% variance)"),
    y = paste0("PC 2 (", pc2_var, "% variance)")
  )+ scale_color_manual(values=c( "#FFD353FF","#FFB242FF","#BB292CFF"), name="i-score 80")+ scale_fill_manual(values=c( "#FFD353FF","#FFB242FF","#BB292CFF"),name="i-score 80")
```

```{r}
x<- coxdata%>% filter(patid %in% coxdata$patid) %>% dplyr::select(patid,starts_with("residual")) %>% arrange(patid) %>% column_to_rownames("patid")
pca_df <- prcomp(x , scale. = F ) 
variance_explained <- summary(pca_df)$importance[2, 1:2] * 100  # Convert to percentage
pc1_var <- round(variance_explained[1], 2)
pc2_var <- round(variance_explained[2], 2)

pca_df <- as.data.frame(pca_df$x)
pca_df$patid <-as.numeric(rownames(x))


pca_df %>%  left_join(coxdata) %>% ggplot(aes(x=PC1,y=PC2)) + geom_point(aes(color=i_score_80_cat)) + theme_classic() + geom_point(
    data = pca_df  %>% right_join(coxdata, ., by = "patid") %>%
      group_by(i_score_80_cat) %>% 
      summarise(  median_PC1 = median(PC1, na.rm = TRUE), median_PC2 = median(PC2, na.rm = TRUE)),
    aes(x = median_PC1, y = median_PC2, fill=i_score_80_cat),  shape=24 ,size=5,color="white"
  ) + ggtitle("PCA residual protein values") +  labs(
    x = paste0("PC 1 (", pc1_var, "% variance)"),
    y = paste0("PC 2 (", pc2_var, "% variance)")
  )+ scale_color_manual(values=c( "#FFD353FF","#FFB242FF","#BB292CFF"), name="i-score 80")+ scale_fill_manual(values=c( "#FFD353FF","#FFB242FF","#BB292CFF"),name="i-score 80")
```

## 4 Correlation Network
```{r}
library(pheatmap)
library(qgraph)
library(Hmisc)   # for rcorr
library(ppcor)
```

create protein networks based on the eGFR residuals of the protein levels ( controlling for eGFR influence)
and also correct the pairwise correlation on the presences/absence of othe proteins (partial correlation)

```{r}
## all samples
x<- coxdata%>% filter(patid %in% coxdata$patid) %>% dplyr::select(patid,starts_with("residual")) %>% arrange(patid) %>% column_to_rownames("patid")
names(x) <- labels$name

```

Partial Correlation Network:
conditional correlation, adjusted by other proteins
```{r}
qg<-qgraph(cor(x), layout = "spring", sampleSize=nrow(x),threshold="fdr",
       labels = colnames(x), theme = "classic", graph="pcor"
      ,color = ifelse(colnames(x) %in% c("CD40","PD-L1","CX3CL1","CD137"), "lightblue", "lightgrey"),legend=T,
  vsize = ifelse(colnames(x) %in% c("CD40","PD-L1","CX3CL1","CD137"), 6, 5))

```

or alternatively ego networks 
showing correlation of the four proteins with other proteins
```{r}
plot_egonetwork <- function(ego){
  adj <- getWmat(qg)


neighbors <- which(adj[ego, ] != 0)
subnodes <- c(ego, names(neighbors))

# plot subnetwork
return(qgraph(adj[subnodes, subnodes], labels = subnodes, theme = "classic", layout = "spring"))
}

plot_egonetwork("CD40")
plot_egonetwork("PD-L1")
plot_egonetwork("CD137")
plot_egonetwork("CX3CL1")

```

connecting ego networks to one
showing a subset of the whole partial correlation network 
```{r}
 adj <- getWmat(qg)
neighbors <- as.data.frame(adj[c("CD40","PD-L1","CX3CL1","CD137"), ] != 0) %>% dplyr::select(where(any))#, -c("CD40","PD-L1","CX3CL1","CD137")) 
subnodes <- names(neighbors)
qgraph(adj[subnodes, subnodes], labels = subnodes,legend=T
       , layout = "spring",
       color = ifelse(colnames(adj[subnodes, subnodes]) %in% c("CD40","PD-L1","CX3CL1","CD137"), "lightblue", "lightgrey"),
  vsize = ifelse(colnames(adj[subnodes, subnodes]) %in% c("CD40","PD-L1","CX3CL1","CD137"), 6, 5))

```



```{r}
centralityTable(qg)
```


```{r}
centralityTable(qg) %>% ggplot(aes(x=value)) + geom_density() + facet_wrap(~measure)
centralityTable(qg) %>% group_by(measure) %>% summarise(mean=mean(value,na.rm=T), median=median(value,na.rm=T))
```


```{r}
centralityTable(qg) %>% filter(measure=="Betweenness") %>% arrange(desc(value))
centralityTable(qg) %>% filter(measure=="Strength") %>% arrange(desc(value))
centralityTable(qg) %>% filter(measure=="Closeness") %>% arrange(desc(value))
centralityTable(qg) %>% filter(measure=="ExpectedInfluence") %>% arrange(desc(value))

```


calculate parital correaltion, get p-values and adjust them (fdr control)
```{r warning=FALSE}
X <- as.matrix(x)  # patients × proteins matrix
P <- ncol(X)

# Initialize matrices
## no lasso regualization, overfitting
pcor_mat <- matrix(NA, nrow = P, ncol = P, dimnames = list(colnames(X), colnames(X)))
pval_mat <- matrix(NA, nrow = P, ncol = P, dimnames = list(colnames(X), colnames(X)))

for (i in 1:(P-1)) {
  for (j in (i+1):P) {
    res <- pcor.test(X[,i], X[,j], X[, -c(i,j)])
    pcor_mat[i,j] <- pcor_mat[j,i] <- res$estimate
    pval_mat[i,j] <- pval_mat[j,i] <- res$p.value
  }
}
diag(pcor_mat) <- 1
diag(pval_mat) <- 0
```
```{r}
pval_vec <- pval_mat[upper.tri(pval_mat)]
fdr_vec <- p.adjust(pval_vec, method = "fdr")

fdr_mat <- matrix(NA, nrow = P, ncol = P, dimnames = list(colnames(X), colnames(X)))
fdr_mat[upper.tri(fdr_mat)] <- fdr_vec
fdr_mat <- t(fdr_mat)
fdr_mat[upper.tri(fdr_mat)] <- fdr_vec
diag(fdr_mat) <- 0
```



```{r}
edges <- which(upper.tri(pcor_mat), arr.ind = TRUE)
edge_df <- data.frame(
  from   = colnames(X)[edges[,1]],
  to     = colnames(X)[edges[,2]],
  weight = abs(pcor_mat[edges]),
  og_weight=pcor_mat[edges],
  fdr    = fdr_mat[edges]
) %>% dplyr::filter(fdr < 0.05)
```

```{r}

Chemokines<-c(
 "CCL7", "CCL2", "CXCL11", "CXCL9", "CXCL1", "CCL4",
"CCL13", "CCL11", "CCL19", "CXCL5", "CCL23", "CCL3",
"CXCL6", "CXCL10", "CCL28", "CCL8", "CCL25", "CX3CL1", "CCL20"
  )
TNF_superfamily<-c(
"TRAIL", "TNFSF10", "TNFSF14", "TNFSF11", "TNF","TNFB", "TNFRSF9","TNFSF11", 
"TNFSF12","TWEAK", "LTA", "TNFRSF11B","OPG", "TRANCE","RANKL"
  )
Interleukins_and_receptors<-c(
"IL7" ,"IL8", "IL6", "IL17C", "IL17A", "IL20RA", "IL18",
"IL10RA", "IL15RA", "IL10RB", "IL18R1", "IL12B", "IL10", "OSM","LIFR","IFNGAMMA"
  )
Growth_factors<-c(
"VEGFA", "GDNF", "TGFB1", "KITLG","SCF", "TGFA","TGFALPHA", "FGF23",
"FGF5", "FGF21", "HGF", "FGF19", "CSF1", "FLT3LG", "NTF3","NT3","LAPTGF_BETA_1"
  )
Immune_costimulation_checkpoint<-c(
"CD274","PDL1", "CD40", "CD244", "CD5", "CD6", "CD8A",	"CDCP1", "SLAMF1"
  )

other <- c("UPA"  ,  "AXIN1" , "CST5"  , "MMP1" ,   "MMP10"  ,"X4EBP1", "SIRT2",  "DNER" ,  "ENRAGE", "CASP8" ,"ST1A1",  "STAMBP", "ADA" )

```

```{r}
node_df <- data.frame(
  name = colnames(pcor_mat)) %>% left_join(labels) %>% mutate(analyte= str_remove( str_to_upper(analyte),"_")) %>% 
  mutate( group = case_when(name %in% Chemokines | analyte %in% Chemokines ~"Chemokines",
                     name %in% TNF_superfamily | analyte %in% TNF_superfamily ~"TNF superfamily",
                    name%in% Interleukins_and_receptors | analyte %in% Interleukins_and_receptors ~ "Interleukins and receptors",
                     name%in% Growth_factors | analyte %in% Growth_factors  ~ "Growth factors",
                     name %in% Immune_costimulation_checkpoint | analyte %in% Immune_costimulation_checkpoint ~ "Immune costimulation checkpoint",
                    name %in% other | analyte %in% other ~ "other"
)) 
node_df %>% filter(is.na(group)) %>% pull(analyte)
```

### figure 5 network plot

from created network, showing subst of the big nwtoqrk concentrating on the significant correlations of our 4 proteins 

```{r}
#"#163343FF",#"#56C0A7FF"

colors <- c(
"i-score"= "#DC1F51", #  "#D41146", #"#990F4B",

"Chemokines"="#EC8FA3FF",
"TNF superfamily"="#FCBA65FF",
"other"= "#FAECCFFF",
"Interleukins and receptors"="#8D7F99FF",
"Growth factors"="#8C9D57FF",
"Immune costimulation checkpoint"="#6FA3B8"
)
```


```{r}
subset_edge_df <- edge_df %>% filter( to %in% c("CD40","PD-L1","CX3CL1","CD137") | from %in% c("CD40","PD-L1","CX3CL1","CD137") ) %>% mutate(og_weight=weight,weight=abs(weight))

#subset_edge_df <- edge_df %>% filter( to %in% union(subset_edge_df$from,subset_edge_df$to) & from %in% union(subset_edge_df$from,subset_edge_df$to) ) %>% mutate(og_weight=weight,weight=abs(weight)) 
subset_node_df <- node_df %>% filter(name %in% union(subset_edge_df$from, subset_edge_df$to)) %>% mutate(group=if_else(name %in% c("CD40","PD-L1","CX3CL1","CD137"), "i-score",group))

subset_g <- graph_from_data_frame(d = subset_edge_df, vertices = subset_node_df, directed = FALSE)

graph<- ggraph(subset_g) + 
  ggraph::geom_edge_link(aes(width = abs(weight), color = og_weight)) +  
  ggraph::geom_node_point(aes(color = group), size = 9) +
  #ggraph::geom_node_text(aes(label = name), repel = F, size = 2) +
  # non-purple labels
  geom_node_text(
    aes(label = name,
        filter = group != "Interleukins and receptors"),
    color = "black",
    repel = FALSE,
    size = 2.2
  ) +

  # purple-group labels
  geom_node_text(
    aes(label = name,
        filter = group == "Interleukins and receptors"),
    color = "white",
    repel = FALSE,
    size = 2.2
  ) +
 # scale_edge_color_gradient2(low =  "white",mid="#F07470", high = "#DC1C13", midpoint = 0.175,name="partial correlation") +
 scale_edge_color_gradient2(low =  "#F8F8FF",mid="#9B9A9C", high = "#3E3B39", midpoint = 0.175,name="partial correlation") +

  scale_edge_width(range = c(0.2, 2),name="partial correlation") + theme_graph() + scale_color_manual(values=colors,name=NULL) 

graph
```


whole network 



```{r}
g <- graph_from_data_frame(d = edge_df, vertices = node_df, directed = FALSE)

graph<- ggraph(g) + 
  ggraph::geom_edge_link(aes(width = abs(weight), color = og_weight)) +  
  ggraph::geom_node_point(aes(color = group), size = 9) +
 # ggraph::geom_node_text(aes(label = name), repel = F, size = 1.8) +
  # non-purple labels
  geom_node_text(
    aes(label = name,
        filter = group != "Interleukins and receptors"),
    color = "black",
    repel = FALSE,
    size = 2.2
  ) +

  # purple-group labels
  geom_node_text(
    aes(label = name,
        filter = group == "Interleukins and receptors"),
    color = "white",
    repel = FALSE,
    size = 2.2
  ) +
  scale_edge_color_gradient2(low =  "blue",mid="white", high = "red", midpoint = 0,name="partial correlation") +
 # scale_edge_color_gradient2(low =  "#F8F8FF",mid="#6D6A6A", high = "#3E3B39", midpoint = 0.175,name="partial correlation") +
  scale_color_manual(values=colors, name=NULL) +
  scale_edge_width(range = c(0.2, 2),name="abs(partial correlation)") + theme_graph() 

graph
```

