---
title: "metadeconfoundR"
output: html_notebook
---
assocation of olink proteins and clinical variables

```{r include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(readxl)
library(paletteer)
library(stringr)
library(tibble)
```

## Data 
```{r}
pat_withoutegfr <- 737

### baseline table: contains all clinical variables at month 0 for each patient 
BL_data_bpoffice <-read.csv2("~/Projects/4cstudy/Tables/BL_data_bpoffice.csv",  sep = ",",header=T) %>% filter(patid != pat_withoutegfr) %>% mutate(across(everything(), ~ type.convert(.x, as.is = TRUE)))

### investigation file from ckd progression, clinical data from all visits month 0 -120 
investigation_file <-read.csv("~/Projects/4cstudy/Tables/investigation_file.csv",  sep = ",",header=T) %>% filter(!is.na(month) & month!="FALSE")%>% mutate(across(everything(), ~ type.convert(.x, as.is = TRUE))) 

### from ckd progression 
# contains the first month where the patient reached one of the endpoints: ckd50%, tx, dialysis
reached_endpoint <- reached_endpoint <- read.csv("~/Projects/4cstudy/Tables/reached_endpoint.csv", sep = ",", header=T) %>% mutate(across(everything(), ~ type.convert(.x, as.is = TRUE)))


### olink data

olink_data <- read.csv("~/Projects/4cstudy/Tables/OLINK_data.csv",  sep = ",",header=T) %>% rename(patid=pat_id) %>%arrange(patid) %>% mutate(across(everything(), ~ type.convert(.x, as.is = TRUE))) %>% filter(patid != 737)

### labels 
labels <- read_excel("~/Projects/4cstudy/ClinicalData/analyte_abbrev_HB.xlsx") 
```


## Iteration Function 
function that run all xx month endpoints
each run looks for the assocation of reachinf the endpoint within xx months and the clinical variables of interest 

the function defines reaching endpoint 
prepares the data input for metadeconfoundR 
runds metadeconfoundr and save output table and heatmap of the detected associations 
```{r}
run_metadeconfoundR_reachep <- function(m, metadata, olinkdata, reachedepdata, protein_prefix="rawscale_", path="C:/Users/rosa/Documents/4CStudy/Plots/mdcR2/", pat_withoutegfr=xxx,  qcutoff=0.05, metanames=NULL){

## include either patients that reachednpoint at month xx or that have an entry at month xx
## hence exclude patients that drop out in between, not counting as survivor
### define reached endpoint within xx months 
epname <- paste0("reachedep", m,"m")
patient_reachep<- reachedepdata %>% filter(month <= m) %>% pull(patid) 

## get patients that have an entry for month xx  
pat_with_visit <- investigation_file  %>% filter(month== as.numeric(m)) %>% distinct(patid) %>% pull(patid)


####################### protein data , subset to patients with data at month xx
olink_input_metadecon <- olinkdata %>% filter(patid != pat_withoutegfr & (patid %in% pat_with_visit  | patid %in% patient_reachep))  %>% arrange(patid) %>% column_to_rownames("patid") %>% select(starts_with(protein_prefix))  %>% mutate(across(everything(), as.numeric))%>%data.frame()

names(olink_input_metadecon)<-str_replace(names(olink_input_metadecon),paste0("^",protein_prefix),"")

#################### meta data
## get all covariates
## country, diagnose, hyper groups as dummies 
meta_input_metadecon_reachedep <- metadata %>% filter(patid != pat_withoutegfr  & (patid %in% pat_with_visit  | patid %in% patient_reachep) )  %>% arrange(patid) %>% select( age,sex,bmisds,egfr,alburie,salb,bicar,hb,phos,diasds,syssds,syssds_above2,diasds_above2, hyper_class, hyper_treat_group,medigr_RAS.inhibitors, medigr_Immunsuppressiva,diagnose2, country, patid) %>% mutate(hyper_class=if_else(hyper_class=="yes",1,0), alburie=log(alburie +0.01)) %>%  mutate(sex_male=if_else(sex=="male",1,0),.keep = "unused", .before = 1 ) %>% 
fastDummies::dummy_columns(. ,select_columns = c("diagnose2", "hyper_treat_group", "country"), remove_selected_columns = T) 

## define reached enpoint
meta_input_metadecon_reachedep <- meta_input_metadecon_reachedep %>% mutate(reachedep=if_else(patid %in% patient_reachep,1,0),.before=1)

### add type of reached endpoint : so that survival is compared with each endpoint 
 meta_input_metadecon_reachedep<-  meta_input_metadecon_reachedep %>% left_join(
   reached_endpoint %>% fastDummies::dummy_columns(. ,select_columns = c( "RRT"), remove_selected_columns = T)%>%  
     select(starts_with("RRT",ignore.case = F) ,patid), by="patid")  %>% mutate(across(starts_with("RRT"), ~ case_when(.x==0 ~NA, is.na(.x) ~ 0, T ~.x)))

## remove this variable
if ("hyper_treat_group_NA" %in% names(meta_input_metadecon_reachedep)){
  meta_input_metadecon_reachedep <- select(meta_input_metadecon_reachedep, -hyper_treat_group_NA )
}
## transform to data frame with rownames
meta_input_metadecon_reachedep<- meta_input_metadecon_reachedep %>% arrange(patid) %>% column_to_rownames("patid") %>% as.data.frame(.)

## set them to numeric variables
meta_input_metadecon_reachedep[,3:12] <- apply(meta_input_metadecon_reachedep[,3:12], 2,as.numeric)


###################################################################################
## run metadeconfoundR
print(paste("running metadeconfoundR", "endpoint", m, "m"))
print(paste(names(table(meta_input_metadecon_reachedep$reachedep)), table(meta_input_metadecon_reachedep$reachedep), sep=":", collapse = " "))


metadecon_output_ep <- MetaDeconfound(
  featureMat =olink_input_metadecon[rownames(olink_input_metadecon)%in%rownames(meta_input_metadecon_reachedep),],
  metaMat = meta_input_metadecon_reachedep,
  returnLong = TRUE,
 logLevel = "ERROR",
 deconfF=c("RRT_Tx","RRT_CKD50","RRT_Dialysis")
)

## pretty names for heatmap
if (is.null(metanames) & ncol(meta_input_metadecon_reachedep)==39){
  
metanames <- data.frame(original=names(meta_input_metadecon_reachedep), new= c(paste0("reached EP ", m, "m"),"sex","age","BMI sds","eGFR","log(alburie +0.01)","albumin","bicarbonat","hemoglobin","phosphorus","dia. sds","sys. sds","sys. sds > 2","dia- sds > 2","hypertension","CAKUT","Glomerulopathies","postAKI/HUS","other disease","Tubint","HTN: treated,normal","HTN: treated,uncontrolled","HTN: untreated,normal","HTN:untreated,uncontrolled","RAS inhibitor","Immunsuppressiva", "Austria","France", "Germany","Italy", "other country", "Poland", "Serbia", "Turkey", "UK","EP:50%loss","EP:Death","EP:Dialysis","EP:Transplantation" ) )
  
}

  
## save outcome
heatmap<- BuildHeatmap(metadecon_output_ep,q_cutoff=qcutoff, metaVariableNames = metanames, featureNames = labels, reOrder = "feat") + theme(axis.text.x = element_text(size=9),axis.text.y = element_text(size=8), plot.title = element_blank() ) + ylab("") + xlab("") + labs( y="",x="",title="", subtitle="FDR: < 0.001 = ***, < 0.01 = **, < 0.05 = *")

#ggsave( paste0(path, "heatmap_reachedendpoint_n", nrow(meta_input_metadecon_reachedep),"_",m,"m.pdf"),heatmap, width=6,height=10) 


#write.table(metadecon_output_ep, paste0(path, "metadeconfoundR_output_reachedendpoint_n", nrow(meta_input_metadecon_reachedep),"_",m,"m.tsv"), row.names=F, quote=F,na="NA", sep="\t")

  
print( paste("number of signif. proteins:" , nrow(metadecon_output_ep %>% filter(metaVariable== "reachedep" & Qs< 0.05 & status == "OK_sd" ) %>% mutate(metaVariable= epname) )))

return(metadecon_output_ep  )
}

```

## Excecute iteration

run function in for loop , from month 12 to 96

saving results
```{r}
metadecon_allresults <- data.frame()
for (i in 1:8){
  res <- run_metadeconfoundR_reachep(m=12*i, metadata = BL_data_bpoffice,reachedepdata=reached_endpoint, olinkdata = olink_data) %>% filter(metaVariable== "reachedep" ) %>% mutate(metaVariable= paste0("reachedep", 12*i,"m"))

  metadecon_allresults <- rbind(metadecon_allresults, res)
}
```

## Summary plot 
summary heatmap, 
taking the columns of each endpoint and adding the assocation with other clinical variables for the endpoint at 12 month
```{r}
res12m <- run_metadeconfoundR_reachep(m=12, reachedepdata=reached_endpoint,metadata = BL_data_bpoffice, olinkdata = olink_data)
```

```{r}
order_mvs<- c(rev(paste0("reachedep", seq(12,96,12), "m")),as.character(unique(res12m$metaVariable)[-1]))

metanames <- data.frame(original=order_mvs, new= c(rev(paste0("reached EP ", seq(12,96,12), "m")),"sex","age","BMI sds","eGFR","log(alburie +0.01)","albumin","bicarbonat","hemoglobin","phosphorus","dia. sds","sys. sds","sys. sds > 2","dia- sds > 2","hypertension","CAKUT","Glomerulopathies","postAKI/HUS","other disease","Tubint","HTN: treated,normal","HTN: treated,uncontrolled","HTN: untreated,normal","HTN:untreated,uncontrolled","RAS inhibitor","Immunsuppressiva", "Austria","France", "Germany","Italy", "other country", "Poland", "Serbia", "Turkey", "UK","EP:50%loss","EP:Dialysis","EP:Transplantation" ))

metadecon_allresults <- metadecon_allresults %>% group_by(feature) %>% filter(any(Qs<0.1))

metadeconR_heatmap_all_covariatesEP12<- BuildHeatmap(
  bind_rows(metadecon_allresults , res12m %>% filter(metaVariable!="reachedep" & feature %in% metadecon_allresults$feature)) %>% mutate(metaVariable=factor(metaVariable, levels=order_mvs))
             ,q_cutoff=0.05, metaVariableNames = metanames, featureNames = labels, reOrder = "feat") + theme(axis.text.x = element_text(size=6.5),axis.text.y = element_text(size=6.5), plot.title = element_blank() )  + labs( y="",x="",title="", subtitle="FDR: < 0.001 = ***, < 0.01 = **, < 0.05 = *")

                        
metadeconR_heatmap_all_covariatesEP12                      
```

```{r}
metadecon_allresults  %>% filter(startsWith(status,"C: R"))
```
### combine with contrasts

```{r}
diagnose_contrasts <- read.table(file="~/Projects/4cstudy/Tables/diagnose_contrast_results.csv", sep = ",",header=T)
country_contrasts <- read.table(file="~/Projects/4cstudy/Tables/country_contrast_results.csv", sep = ",",header=T) 

diagnose_contrasts<- diagnose_contrasts %>% rename(feature=analyte, metaVariable=contrast,Ps=p.value, Ds=effect_size) %>% mutate(status=if_else(Ps<0.05,"OK_sd","NS"),Qs=Ps) %>% dplyr::select(feature,metaVariable,Ps,Qs,Ds,status)

country_contrasts<- country_contrasts %>% rename(feature=analyte, metaVariable=contrast,Ps=p.value, Ds=effect_size) %>% mutate(status=if_else(Ps<0.05,"OK_sd","NS"),Qs=Ps) %>% dplyr::select(feature,metaVariable,Ps,Qs,Ds,status)
```
```{r}
metadata_0m<- BL_data_bpoffice %>% filter(patid != pat_withoutegfr)  %>% arrange(patid) %>% select( age,sex,bmisds,egfr,alburie,salb,bicar,hb,phos,diasds,syssds,syssds_above2,diasds_above2, hyper_class, hyper_treat_group,medigr_RAS.inhibitors, medigr_Immunsuppressiva,diagnose2, country, patid) %>% mutate(hyper_class=if_else(hyper_class=="yes",1,0), alburie=log(alburie +0.01)) %>%  mutate(sex_male=if_else(sex=="male",1,0),.keep = "unused", .before = 1 ) %>% 
fastDummies::dummy_columns(. ,select_columns = c("diagnose2", "hyper_treat_group", "country"), remove_selected_columns = T) 

metadata_0m <- metadata_0m %>% arrange(patid) %>% column_to_rownames("patid")

olink_0m <- olink_data %>% filter(patid != pat_withoutegfr)  %>% arrange(patid) %>% column_to_rownames("patid") %>% select(starts_with("rawscale"))  %>% mutate(across(everything(), as.numeric))%>%data.frame()

names(olink_0m)<-str_remove(names(olink_0m),"rawscale_")

res0m <- MetaDeconfound(
  featureMat =olink_0m[rownames(olink_0m)%in%rownames(metadata_0m),],
  metaMat = metadata_0m,
  returnLong = TRUE,
 logLevel = "ERROR"
)
```


```{r}
metanames <- data.frame(original=unique(res0m$metaVariable), new= c("sex (male)","age","BMI sds","eGFR","UACR","albumin","bicarbonat","hemoglobin","phosphorus","dia. sds","sys. sds","sys. sds > 2","dia. sds > 2","hypertension","RAS inhibitor","Immunsuppressiva","CAKUT","Glomerulopathies","postAKI/HUS","other disease","Tubint","HTN: treated,normal","HTN: treated,uncontrolled","HTN: untreated,normal","HTN:untreated,uncontrolled","HTN:NA","Austria","France", "Germany","Italy", "other country", "Poland", "Serbia", "Turkey", "UK" )) %>% rbind(data.frame(original=unique(diagnose_contrasts$metaVariable),new=unique(diagnose_contrasts$metaVariable))) %>%
  rbind(data.frame(original=unique(country_contrasts$metaVariable),new=unique(country_contrasts$metaVariable)))

BuildHeatmap(res0m %>% mutate(Ds=scale(Ds))%>% filter(! startsWith(as.character(metaVariable),"country") & (!startsWith(as.character(metaVariable),"diagnose")) ) %>% rbind(diagnose_contrasts) %>% rbind(country_contrasts) , 
              metaVariableNames = metanames, featureNames = labels,
             reOrder="feat",q_cutoff = 0.05 ) +labs( y="",x="",title="", subtitle="FDR/Tukey p-val: < 0.001 = ***, < 0.01 = **, < 0.05 = *")
```

