---
title: "Survival Analysis Preparation"
output: html_notebook
---

this scripts contains the calculation of survival time (reaching the endpoint), the calculation of the inflammation score, and plotting of Kaplan-Meier Curves for the different inflammation score groups as well as the distirbution of eGFR and UACR in those groups. 


```{r libraries, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(readxl)
library(paletteer)
library(tibble)
library(stringr)
library(survival)
library(purrr)
library(ggpubr)
library(survminer)
library(broom)
library(gridExtra)
library(viridisLite)
library(ggstats)
library(matrixStats)
library(purrr)
library(corrplot)
```

```{r}
source("~/Projects/4cstudy/code_upload/survival_analysis_functions.R")
```

## 1 Data
clinical data: 
```{r}
pat_withoutegfr <- xxx

## clinical data at study entrance (egfr, proteinuria,...)
BL_data_bpoffice <-read.csv2("~/Projects/4cstudy/Tables/BL_data_bpoffice.csv",  sep = ",",header=T) %>% filter(patid != pat_withoutegfr) %>% mutate(across(everything(), ~ type.convert(.x, as.is = TRUE)))

### other patient info, non related to a specific visit
patients_file <- read_excel("~/Projects/4cstudy/ClinicalData/patients_dataset_4C_for_OLINK.xlsx")

### investigation file from ckd progression, clinical data from all visits month 0 -120 
investigation_file <-read.csv("~/Projects/4cstudy/Tables/investigation_file.csv",  sep = ",",header=T) %>% filter(!is.na(month) & month!="FALSE")%>% mutate(across(everything(), ~ type.convert(.x, as.is = TRUE)))  #%>% 
### from ckd progression 

# contains the first visit where the patient reached one of the endpoints: ckd50%, tx, dialysis
reached_endpoint <- read.csv("~/Projects/4cstudy/Tables/reached_endpoint.csv", sep = ",", header=T) %>% mutate(across(everything(), ~ type.convert(.x, as.is = TRUE)))

### olink protein values 
olink_data <- read.csv("~/Projects/4cstudy/Tables/OLINK_data.csv",  sep = ",",header=T) %>% arrange(pat_id) %>% rename(patid=pat_id)%>% mutate(across(everything(), ~ type.convert(.x, as.is = TRUE)))
olink_data <- olink_data  %>% select(patid, starts_with("rawscale")) %>% filter(patid != pat_withoutegfr) %>% mutate(across(everything(), as.numeric))
names(olink_data) <- str_replace(names(olink_data), "rawscale_", "native_")

```


## 2 Calculate Survival Time

### Death
little modicfation for the composite endpoint: adding three patients that died from drop outs to group of reaching endpoint
getting patients id and age at death
```{r}
death_patients <- setdiff(patients_file %>% filter(doreason =="Death") %>% pull(patid) ,reached_endpoint$patid )
death_age <- patients_file %>% filter(patid %in% death_patients) %>% pull(age_in_years_at_deathdate,patid)

patients_file %>% filter(patid %in% death_patients) %>% select(patid,contains("do"))
investigation_file %>% filter(patid %in% death_patients)

reached_endpoint<- reached_endpoint %>% mutate(across(c(vdate,birthdate),as.Date)) %>% mutate(across(contains("age_"),as.numeric))%>% bind_rows(investigation_file %>% mutate(across(c(vdate,birthdate),as.Date)) %>% group_by(patid) %>% filter(patid %in% death_patients & month==max(month)) %>% mutate(RRT="Death") ) 
```

### RRT
get patients ids for the ones on RRT (Tx or Dialysis)
```{r}
dialysis_pats<- reached_endpoint %>% filter(RRT=="Dialysis") %>% pull(patid)
tx_pats<- reached_endpoint %>% filter(RRT=="Tx") %>% pull(patid)

egfrloss_pats_month <- reached_endpoint %>% filter(RRT=="CKD50") %>% select(patid,epm=month) #epm: endpoint month
```

### 50% eGFR loss linear interpolation

for 50% egfr loss -\> linear interpolation taken all egfr values before
and at month where 50% egfr loss is reached , based on this determine
age where loss is reached

```{r}
age_at_50egfr_df <- investigation_file %>% filter(patid %in% egfrloss_pats_month$patid) %>% left_join(egfrloss_pats_month,by="patid") %>% group_by(patid) %>% filter(month <= epm) %>% arrange(month)  %>%
  summarise( age_at_50egfr= approx(x=egfr,y=age_in_years_at_vdate, xout=0.5*first(egfr))$y)
```

### create survival data

survival time, time frame in years between Baseline and endpoint is
reached 1: reached, 0: censored (last available month)

```{r}
## add age at study entrance BL_age
survival_data <- investigation_file %>% left_join(BL_data_bpoffice %>% mutate(BL_age= as.numeric(age_in_years_at_vdate) )%>% select(patid,BL_age), by="patid") 
## add age_at_50egfr as column to data frame
survival_data<- survival_data %>% left_join(age_at_50egfr_df,by="patid") 
## calculate survival time as the age passed from study entrance (BL_age) to tx/rrt/death/50%loss or drop out
survival_data <- survival_data %>% group_by(patid)%>% 
mutate(survival_time = case_when(patid %in% dialysis_pats  ~ age_in_years_at_diadate -BL_age, 
                              patid %in% tx_pats ~ age_in_years_at_tpldate-BL_age,
                              as.character(patid) %in% death_patients ~ death_age[as.character(patid)] -BL_age,
                              patid %in% age_at_50egfr_df$patid ~ as.numeric(age_at_50egfr)-BL_age,
                              ## for drop outs , censored data, get age of last visit
                              !(patid %in% reached_endpoint$patid) ~ age_in_years_at_vdate[month==max(month)]-BL_age )) 

## in case no age is given, remove patients without survival time and take the minimum survival time for each patient
survival_data <- survival_data %>% filter(!is.na(survival_time)) %>% slice_min(survival_time) %>% distinct(patid,survival_time) 
### add binary reached endpoint 
survival_data <- survival_data %>%mutate(reachep= if_else(patid %in% reached_endpoint$patid ,1,0)) 
```
```{r}
## remove patients with only one entrance and being censored
only_one_entrance_ids <- investigation_file %>% group_by(patid) %>% filter(n()==1) %>% pull(patid)

survival_data <- survival_data %>% filter(!(patid %in% only_one_entrance_ids & reachep==0) )
```

add clinical at study entrance to survival data
```{r}
survival_clinical_data <- survival_data %>% select(patid,survival_time,reachep) %>% left_join(BL_data_bpoffice,by="patid")

```

## 3 Calculate Inflammation Score

first step: linear model protein ~ eGFR + alburie, get residuals from this model
protein levels not explained by eGFR or alburie 

```{r}
olink_data <- olink_data %>% filter(patid %in% survival_clinical_data$patid ) %>% arrange(patid) 

residuals_proteins <- sapply(names(olink_data %>% select(starts_with("native"))), function(x){

      form <- formula(paste(x, "~ egfr + log(alburie +0.01)" ))
     model <- lm(form, data= left_join(survival_clinical_data %>% arrange(patid), olink_data %>% select(patid, {{x}}) ,by="patid" )%>% arrange(patid),
                 na.action = "na.exclude")
    
     return(residuals(model))}) %>%
    ## add patient ids , for all with eGFR and alburie information
    as.data.frame() %>% mutate(patid =olink_data$patid) %>% relocate(patid) %>% filter(!is.na(native_il8))

## change protein names
names(residuals_proteins) <- str_replace(names(residuals_proteins),"native_", "residual_")
```

second step: labels residuals as "unsual" high (not explained by ckd status) if residuals is above the xxth percentile (50th-90th)
```{r}
## label residuals as high or not high for each cut off
 for (cutoff in seq(0.5, 0.9, 0.1)) {
  residuals_proteins <- residuals_proteins %>% mutate(across(
    starts_with("residual_"),
    ~ ifelse(. >= quantile(., cutoff, na.rm = TRUE), 1, 0),
    .names = paste0("high_", cutoff*100, "_{.col}")
  ))
 }
```

third step: 
i-score for the five proteins in protein_set
sum up all resifuals above threshold = i-score
i_score= for the 80th cutoff 
but also for the other cutoffs

```{r}
protein_set <- c("tnfrsf9","pd_l1","cx3cl1","cd40")
setnames<- c("CD137","PD_L1","CX3CL1","CD40")
```

```{r}
residuals_proteins<- residuals_proteins %>% group_by(patid) %>% mutate(i_score= as.character(sum(c_across(paste0("high_80_residual_",protein_set)), na.rm = TRUE))) %>% ungroup()

iscores <- reduce(seq(0.5, 0.9, 0.1), function(df, cutoff) {
  cutoff_perc=cutoff*100
  colnames <- paste0(paste0("high_", cutoff_perc, "_residual_"),protein_set)
  score_col <- paste0("i_score_", cutoff_perc)
  
  df %>%
    group_by(patid) %>%
    mutate(!!score_col := as.character(sum(c_across(colnames), na.rm = TRUE))) %>%
    ungroup()
}, .init = residuals_proteins)

#write.table(iscores, "~/Projects/4cstudy/Tables/residuals_iscores.csv" , row.names=F, quote=F,na="NA", sep=",")
```


## 4 data merge 
```{r}
## add survival time, reachep, and subset of clinical data of interest
surv_data <- survival_clinical_data %>% select(patid,survival_time, reachep,egfr,ckd1, sex,BLage=age_in_years_at_vdate,bmisds,diagnose2,alburie,salb,hb,bicar,phos,syssds,diasds,syssds_above2,diasds_above2,hyper_class,hyper_treat_group) 

### add residuals of the protein set of interest : residuals, above cutoff, i-score (80th cut-off)
surv_data <- surv_data %>% right_join(residuals_proteins,by="patid") 

## add the other i-scores with cut off 50-90
surv_data <- surv_data %>% left_join(iscores %>% select(patid, starts_with("i_score_")),by="patid")

## set types
surv_data <- surv_data %>% mutate(i_score=as.numeric(i_score), diagnose2=as.factor(diagnose2), hyper_treat_group=as.factor(hyper_treat_group), hyper_class=as.factor(hyper_class),sex=as.factor(sex))

## input for cox regression
#write.table(surv_data, "~/Projects/4cstudy/Tables/survival_data.csv" , row.names=F, quote=F,na="NA", sep=",")
```

## 5 Kaplan Meier Curves
- relation of i-score and eGFR/alburie 
- kaplan meier curve for each i-score grp: 0 ,1,..,4

### run for protein set of interest
```{r}
protein_set <- c("tnfrsf9","pd_l1","cx3cl1","cd40")
setnames<- c("CD137","PD_L1","CX3CL1","CD40")
```

for cutoff 50 and 80
KMS:
```{r}
plot_kaplan_meier_iscore(data=surv_data, proteinset = c("tnfrsf9","pd_l1","cx3cl1","cd40"), proteinnames =c("CD137","PD_L1","CX3CL1","CD40"),category = T,cutoff="80",prefix="high_80_residual_",color= c(   "#FFD353FF","#FFB242FF","#BB292CFF") ,fontsize=3.5,survline = "hv")

plot_kaplan_meier_iscore(data=surv_data, proteinset = c("tnfrsf9","pd_l1","cx3cl1","cd40"), proteinnames =c("CD137","PD_L1","CX3CL1","CD40"),category = T,cutoff="50",prefix="high_50_residual_",color= c(  "#FFD353FF","#FFB242FF","#BB292CFF") )
```

```{r}
## number of drop outs 
surv_data %>% ungroup() %>% mutate(iscore_cat= case_when(i_score==0~"0", i_score>0 & i_score<3 ~"1-2",T~ "3-4")) %>% dplyr::count(iscore_cat,reachep) %>% group_by(iscore_cat) %>% mutate(percentage=n/sum(n)) %>% write.table("~/Projects/4cstudy/Tables/dropout_numbers_iscore.csv", sep = ",", row.names = F)
```



Boxplot eGFR and alburie
```{r}
plot_boxplot_iscore(data=surv_data,proteinset=c("tnfrsf9","pd_l1","cx3cl1","cd40"),proteinnames=c("CD137","PD_L1","CX3CL1","CD40"),prefix="high_80_residual_",cutoff = "80",xlabel=0.8,psize=3.3)

plot_boxplot_iscore(data=surv_data,proteinset=c("tnfrsf9","pd_l1","cx3cl1","cd40"),proteinnames=c("CD137","PD_L1","CX3CL1","CD40"),prefix="high_50_residual_",cutoff = "50",xlabel=0.8,psize=3.3)

```






KM curves for reaching KRT (splitting endpoint)
```{r}
RRTsurv_data <- surv_data %>% left_join(reached_endpoint %>% dplyr::select(patid,RRT)) %>% filter(RRT=="CKD50"| reachep==0)

plot_kaplan_meier_iscore(data=RRTsurv_data, proteinset = c("tnfrsf9","pd_l1","cx3cl1","cd40"), proteinnames =c("CD137","PD_L1","CX3CL1","CD40"),timelimit = c(0,3),category =T,cutoff="80",prefix="high_80_residual_",color= c( "#FFD353FF","#FFB242FF","#BB292CFF") ,fontsize=3.5)

plot_boxplot_iscore(data=RRTsurv_data,proteinset=c("tnfrsf9","pd_l1","cx3cl1","cd40"),proteinnames=c("CD137","PD_L1","CX3CL1","CD40"),prefix="high_80_residual_",cutoff = "80",xlabel=0.8,psize=3.3)

```
```{r}
reached_endpoint %>% inner_join(surv_data %>% dplyr::select(patid,i_score)) %>%dplyr::count(RRT, i_score)
```



## 6 Residual Visualization

visualizing residuals and cut-offs for the inflammation score definition

```{r}
names<- c("CD137","PD_L1","CX3CL1","CD40")
plist<- list()
for (i in 1:4){
  p<-protein_set[i]
  name<- names[i]
 plist[[p]]<- plot_residual(p,name,savepath="~/Projects/4cstudy/Plots/SurvivalAnalysis/Figure4/residual_plots/")
}

plist
```
# 7 Correlation Plot 

```{r}
cor_df <- BL_data_bpoffice %>% mutate(alburie=log(alburie+0.01)) %>% dplyr::select(patid,egfr,alburie,crp,salb,phos,hb,diasds,syssds) %>% right_join(olink_data%>% select(CD40=native_cd40,CD137=native_tnfrsf9,PD_L1=native_pd_l1,CX3CL1=native_cx3cl1,patid),.) %>% filter(patid %in% surv_data$patid) %>% select(-patid) 


cor_mat <- cor(cor_df,use="complete.obs") 

testRes <- cor.mtest(cor_df, conf.level = 0.95,use="complete.obs")
```


```{r}
corrplot(cor_mat,
         p.mat = testRes$p,
         diag = FALSE,
         order = "FPC",
         type = "lower",
         sig.level = c(0.001, 0.01, 0.05),
         pch.cex = 0.8,
         insig = 'label_sig',
         pch.col = 'grey20',
         tl.col = "black")

# Allow new plot on top
par(new = TRUE)
plot(0, type = "n", axes = FALSE, xlab = "", ylab = "")

# Add the legend
legend("topright",
       legend = c("*** p < 0.001", "** p < 0.01", "* p < 0.05"),
       bty = "n",
       cex = 0.8)
```

# 8 Supplement Figure

```{r warning=FALSE}
source("~/Projects/4cstudy/code_upload/survival_analysis_functions.R")

names<- c("CD137","PD_L1","CX3CL1","CD40")
plist<- list()
for (i in 1:4){
  p<-protein_set[i]
  name<- names[i]
 plist[[p]]<- plot_npx_survtime(p,name,savepath="~/Projects/4cstudy/Plots/SurvivalAnalysis/Figure4/residual_plots/")
}

plist
```

