---
title: "LASSO Regression"
output: html_notebook
---

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(tibble)
library(metadeconfoundR)
library(ggpubr)
library(readxl)
library(paletteer)
library(stringr)
library(glmnet)
library(mlr)
```

## Data Input
```{r message=FALSE, warning=FALSE}
### baseline table: contains all clinical variables at month 0 for each patient 
BL_data_bpoffice <-read.csv2("~/Documents/4CStudy/Tables/BL_data_bpoffice.tsv",  sep = "\t",header=T)

### investigation file from ckd progression, clinical data from all visits month 0 -120 
investigation_file <-read.csv2("~/Documents/4CStudy/Tables/investigation_file.tsv",  sep = "\t",header=T) %>% filter(!is.na(month) & month!="FALSE")

### from ckd progression 
# contains the first month where the patient reached one of the endpoints: ckd50%, tx, dialysis
reached_endpoint <- read.csv("~/Documents/4CStudy/Tables/reached_endpoint.csv", sep = ",", header=T) 

### olink data
olink_data <- read.csv2("~/Documents/4CStudy/Tables/OLINK_data.tsv",  sep = "\t",header=T) %>% arrange(pat_id) %>% rename(patid=pat_id) %>% mutate(across(everything(), as.numeric))


pat_withoutegfr <- "xx"

### labels 
labels <- read_excel("C:/Users/rosa/Documents/4CStudy/ClinicalData/analyte_abbrev_HB.xlsx") 
```


## data preparation


defining predictor categories
- demographics: age,sex bmi
- ckd related: ckd stage, alburie, eGFR
- diagnosis
- bloodpressure variables: systolic, diastolic, hypertension category (yes/no, treated/untreated)
- site: country of center
```{r}
demographic_variables <- c("age", "sex",  "bmisds")
ckd_variables <- c("egfr", "ckd1" , "alburie")
site_variables <- c("country","center") ## center -> overfitting
diagnose_variables <- c("diagnose2")
bloodpressure_variables<- c("syssds","diasds","syssds_above2","diasds_above2", "hyper_class", "hyper_treat_group")

variables <- c(demographic_variables, ckd_variables,site_variables,diagnose_variables,  bloodpressure_variables)

### variable fit 
## filter out without egfr and ckd (n=10) and hypertension group and olink sample
patid_withckd1_hyper <- BL_data_bpoffice%>% filter(patid %in% olink_data$patid & !is.na(ckd1) & !is.na(hyper_treat_group)) %>% pull(patid)
## make dummys from categories (hyper/bp group and class, sex,...)
BL_data_bpoffice_imputed <- BL_data_bpoffice%>% filter(patid %in% patid_withckd1_hyper) %>% arrange(patid)%>% dplyr::select(patid, variables) %>% fastDummies::dummy_cols()

## impute sys and dia with median, and other missing values
BL_data_bpoffice_imputed<- impute(BL_data_bpoffice_imputed %>% dplyr::select(where(is.numeric)), classes=list(numeric=imputeMedian()))$data
```

### Run Lasso 

```{r}
## partially taken the code from  https://doi.org/10.1038/s41591-022-02014-8

run_lasso<- function(data, predictors, start,end){
  set.seed(100)
  res <- data.frame(matrix(ncol=4,nrow=0))
  j<-1
  for( i in start:end){
    y <- data.matrix(data[,i])
    analyte<- names(data)[i]
    ## first cross-fold validation, to find best lambda
    cv.model <- cv.glmnet(predictors, y, alpha=1, nfolds=10, type.measure = "mse")
    
    best_lambda <- cv.model$lambda.min
    
    # fit model with best lambda
    best_model <- glmnet(predictors, y, alpha = 1, lambda = best_lambda)
    
    # taken from publication https://doi.org/10.1038/s41591-022-02014-8:
    if(length(which(as.vector(coef(cv.model, s='lambda.min'))!=0))==1){
          fit=c(analyte, "0","0","1")
        }else{
          selected<-which(as.vector(coef(cv.model, s='lambda.min'))!=0)[-1]-1
      
          selected_predictors=as.data.frame(predictors[,selected])
          selected_predictors$analyte <- as.vector(data.matrix(data[,i]))
          # build model with the best selected predictors
          fit=summary(lm(analyte~.,selected_predictors))
        
          # taken from publication:
          # save adjusted r-square for this model
          fit=c(analyte,ncol(selected_predictors)-1,fit$adj.r.squared,
                pf(fit$fstatistic[1],fit$fstatistic[2],fit$fstatistic[3],lower.tail = F))
        }
    res[j,]<- fit
    j <- j+1
  }
  
  names(res) <- c("analyte", "num_predictors", "adjRsqr", "p_val")
  return(res)
}
```


run each lasso on set of predictors (demographics, bp, ckd,...)
```{r}
# subset to patients with available ckd and hypertension information 
olink_input_lasso <- olink_data%>% filter(patid %in% patid_withckd1_hyper) %>% arrange(patid)

############ demo
lasso_analyte_res_demo <- run_lasso(data = olink_input_lasso,   
                                  predictors =   data.matrix(BL_data_bpoffice_imputed %>% dplyr::select(starts_with("sex"), age,bmisds)), 82,158)

############### bp 
lasso_analyte_res_bp <- run_lasso(data = olink_input_lasso,
                             predictors=  data.matrix(BL_data_bpoffice_imputed %>% dplyr::select(syssds,diasds,syssds_above2,diasds_above2, starts_with("hyper"))), 82,158)


############# ckd
lasso_analyte_res_ckd <- run_lasso(data = olink_input_lasso, 
                                  predictors =  data.matrix(BL_data_bpoffice_imputed %>% dplyr::select(egfr, starts_with("ckd"), alburie)), 82,158)

######## site/country
lasso_analyte_res_country <- run_lasso(data = olink_input_lasso, 
                                  predictors =  data.matrix(BL_data_bpoffice_imputed %>% dplyr::select( starts_with("country"))), 82,158)


######## diagnose
lasso_analyte_res_diagnose <- run_lasso(data = olink_input_lasso, 
                                  predictors =  data.matrix(BL_data_bpoffice_imputed %>% dplyr::select( starts_with("diagnose2"))), 82,158)

```


put the result alltogether, run fdr correction 
```{r}
significant_variation_expl <- lasso_analyte_res_demo %>% mutate(class="demo")   %>% 
  rbind(lasso_analyte_res_ckd %>% mutate(class="ckd")) %>% 
  rbind(lasso_analyte_res_bp%>% mutate(class="bp")) %>% 
 rbind(lasso_analyte_res_site %>% mutate(class="center")) %>% 
  rbind(lasso_analyte_res_country %>% mutate(class="country")) %>%
  rbind(lasso_analyte_res_diagnose %>% mutate(class="diagnose")) %>% 
  mutate(p_val=as.numeric(p_val), adjRsqr=as.numeric(adjRsqr), analyte=str_replace(analyte,"^rawscale_",""))  %>% mutate(class=factor(class,levels=rev(c( "center","country","diagnose","bp", "demo","ckd"))))
```

#### Visualization
visuliaze the variance explained for each predictor class , only with FDR-survived models
```{r}
library(paletteer)

significant_variation_expl %>% mutate(fdr= p.adjust(p_val,method="fdr")) %>% filter(fdr <0.05) %>% left_join( labels, by="analyte") %>% 
  ggplot(aes(x=reorder(name,-adjRsqr, FUN = sum), y=adjRsqr, fill=class )) +
  geom_bar(position="stack", stat="identity") + theme_classic()+
  ylab("adj. R²") + xlab("") + 
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  scale_fill_paletteer_d("LaCroixColoR::MelonPomelo", name="category")
```